\chapter{Exception and Interrupt Handling}
\label{ch:exceptions}

\section{Overview}

An exception is an event that causes the program flow to be diverted to an address stored in a predefined vector.

Some are directly caused by an instruction, some are a side effect of an instruction and some are asynchronously caused
by external hardware.

They are managed by the exception coprocessor, which will take control of the processor when there is a pending
exception.

\section{Exception Types}

\subsection{Faults}

Faults are priority 7 exceptions raised by the CPU itself in response to error conditions or debug events. They use
vectors in the 0x01--0x0F range. Unlike hardware and software exceptions, faults cannot be masked or deferred: when a
fault condition is detected, the fault is dispatched at the next instruction boundary regardless of the current
exception level.

The T bit (TraceMode) in the status register is always cleared when entering any exception handler, including fault
handlers. This ensures that trace faults do not fire inside handlers. RETE restores the saved status register, so if T
was set before the exception, tracing resumes after RETE returns.

\subsubsection{Fault Categories}

Not all faults behave the same way with respect to the faulting instruction and the RETE return address. There are
three categories:

\begin{description}
	\item[Retryable faults] The instruction that triggered the fault is cancelled before it takes effect. The link register
	      stores the address of the faulting instruction. RETE returns to that instruction so it can be retried (possibly after
	      the fault handler has corrected the conditions that caused the fault, e.g.\ mapping a page in response to a bus fault).

	      \textit{Bus Fault, Bus Protection Fault, Alignment Fault, Segment Overflow Fault, Invalid Opcode Fault,
		      Privilege Violation Fault.}

	\item[Post-instruction faults] The instruction completes normally and its effects are committed before the fault fires. The
	      link register stores the address of the \textit{next} instruction. RETE advances to that instruction; the traced
	      instruction is not re-executed.

	      \textit{Instruction Trace Fault.}

	\item[Exception dispatch faults] Raised during the exception handling machinery itself, not during normal instruction
	      execution. The link register stores the program counter at the point the fault was detected within the exception
	      handler. Recovery may require careful state management; these faults often indicate a hardware misconfiguration or a
	      programming error in an exception handler.

	      \textit{Level Five Hardware Exception Conflict, Double Fault.}
\end{description}

\subsubsection{Fault Descriptions}

\begin{itemize}
	\item \textbf{Bus Fault} \textit{(Retryable)} — Raised when an external device asserts the bus error CPU pin.

	      This could happen, for example, if an unmapped address is presented by the CPU and another chip detects this and raises
	      an error. It could also be used to implement virtual memory: the instruction is aborted and the program counter is not
	      incremented, so the address can be re-calculated and the operation retried.

	\item \textbf{Bus Protection Fault} \textit{(Retryable)} — Raised when an external device asserts the bus
	      protection error CPU pin.

	      This indicates that the requested address and device are valid, but the operation is disallowed. Examples include
	      writing to ROM or accessing a protected memory region while the CPU is in protected mode.

	\item \textbf{Alignment Fault} \textit{(Retryable)} — Raised when fetching instructions if the program counter
	      contains an odd word address.

	      This is to simplify fetching: the second word of an instruction is always at \texttt{pl | 0x1}, which ensures that
	      instructions can never overflow a segment boundary.

	\item \textbf{Segment Overflow Fault} \textit{(Retryable)} — Raised when a computed address would fall outside
	      the current segment.

	      Only raised if \texttt{SR.A} (Trap on Address Overflow) is set, since some programs intentionally rely on address
	      wrap-around. Also raised if the program counter overflows; the captured bus address and bus access type in the fault
	      metadata register help identify where the overflow occurred.

	\item \textbf{Invalid Opcode Fault} \textit{(Retryable)} — Raised when a coprocessor call targets a non-existent
	      coprocessor or requests an operation that the selected coprocessor does not implement.

	      Can be used for forward compatibility: programs written for a future co-processor (e.g.\ a floating-point unit) will
	      trap on older hardware, and the handler can emulate the operation in software.

	      Normal processing-unit encodings outside the documented instruction set are reserved or architecturally undefined. They
	      are not required to raise an invalid-opcode fault.

	\item \textbf{Privilege Violation Fault} \textit{(Retryable)} — Raised when a privileged operation is attempted
	      while not in supervisor mode:
	      \begin{enumerate}
		      \item Directly writing to the high word of any address register
		      \item Address-register-pair write-back that would change a high address-register word
		      \item Executing a supervisor-only coprocessor operation
		      \item Triggering a software exception with a vector below 0x60
	      \end{enumerate}

	      In protected mode, writes to the privileged byte of the status register are masked rather than faulting: the low byte
	      is updated, while privileged status bits retain their previous values.

	\item \textbf{Instruction Trace Fault} \textit{(Post-instruction)} — Raised after every instruction completes
	      when the \texttt{TraceMode} SR bit is set.

	      Used for single-step debugging. Because the fault fires after the instruction commits, RETE returns to the instruction
	      \textit{following} the traced one. The T bit is cleared on entry to the fault handler (as with all exceptions), so the
	      handler itself is not traced. RETE restores the saved SR, which re-enables tracing unless the handler explicitly clears
	      T from the saved status register via \texttt{ETTR} before returning.

	\item \textbf{Level Five Hardware Exception Conflict} \textit{(Exception dispatch)} — Raised when a level five
	      hardware exception arrives while one is already being serviced.

	      The windowed link register scheme has no spare register above level 6, so a second L5 cannot be handled normally.
	      Rather than silently dropping it (which could mask a hardware misconfiguration), the CPU raises this fault. Level 5
	      interrupts should be treated as NMIs.

	\item \textbf{Double Fault} \textit{(Exception dispatch)} — Raised when a fault occurs while another fault is
	      already being handled.

	      When the CPU is already at fault level (priority 7) and another fault occurs, there is no additional banked link
	      register available. The fault link register (register 6) is overwritten with the new return address, losing the
	      original fault's return address. The fault metadata register (register 7) is also overwritten; it encodes a
	      double-fault flag along with both the current and original fault types.

	      The CPU jumps to vector 0x08. This is typically a last-chance handler that dumps registers and resets. If a fault
	      occurs while handling a double fault, another double fault is raised, continuously overwriting the link registers.

\end{itemize}

\subsection{Hardware Exceptions (Interrupts)}

A hardware exception occurs when interrupt pins are signalled by external hardware. These are also referred to as
"interrupts". When a hardware exception occurs the current instruction is completed, and control is transferred to
different vectors based on the interrupt "priority". The vectors for hardware exceptions are in the 0x10-0x50 range.

\subsection{User Exceptions (Traps)}

A user exception occurs when the program executes an EXCP instruction that signals the exception coprocessor that it
should trigger an exception. They are usually used by programs running in protected mode to switch to supervisor mode
in a controlled way. They are also referred to as "traps". The vectors for user exceptions are in the 0x60-0xFF range.

\section{Exception Registers}

The SIRC-1 exception coprocessor maintains several special-purpose registers for exception handling.

\subsection{Link Registers}

The exception unit uses banked link registers to store the processor state when an exception occurs. There are eight
link registers in total, indexed 0-7:

\begin{itemize}
	\item Register 0: Software exceptions (traps)
	\item Registers 1-5: Hardware exceptions (interrupts), one for each priority level
	\item Register 6: Faults (abort exceptions)
	\item Register 7: Fault metadata (stores bus address and bus access type information, see section
	      \ref{sec:fault-metadata-register-format} for the format)
\end{itemize}

Each link register stores two pieces of information:

\begin{itemize}
	\item \textbf{Return Address:} The 32-bit address to return to when the exception handler completes
	\item \textbf{Return Status Register:} The 16-bit status register value at the time of the exception
\end{itemize}

When an exception occurs, the appropriate link register (determined by the exception priority level) is populated with
the current program counter and status register values. For abort exceptions (faults), the return address is the
address of the faulting instruction so it can be retried. For regular exceptions (hardware and software exceptions),
the return address is the address after the current instruction.

\subsubsection{Link Register Preservation}

\textbf{Important:} Link registers can be overwritten when a fault occurs during the handling of another fault (a double fault).
In this case, the fault link register (register 6) and the fault metadata register (register 7) is overwritten.
This means the original return address is lost.

A valid way of handling this is just to declare the CPU state as invalid, do some last chance error handling (e.g.
dumping the registers) and reset.

For robust fault handling that can recover from double faults, fault handlers should save the link register contents
early in the handler:

\begin{lstlisting}
:fault_handler
    ; Immediately save link register to memory before doing anything
    ; that might trigger another fault
    ETFR #6                   ; Save level 6 exception registers to a and r7
    STOR -(#0, s), r7
    STOR -(#0, s), ah
    STOR -(#0, s), al
    ETFR #7                   ; Save level 7 (metadata) exception registers to a and r7
    STOR -(#0, s), r7
    STOR -(#0, s), ah
    STOR -(#0, s), al

    ; Now safe to perform operations that might fault
    ; ...

    ; Restore and return
    LOAD al, (#0, s)+
    LOAD ah, (#0, s)+
    LOAD r7, (#0, s)+
    ETTR #7                  ; Restore level 7 (metadata) exception registers from a and r7
    LOAD al, (#0, s)+
    LOAD ah, (#0, s)+
    LOAD r7, (#0, s)+
    ETTR #6                  ; Restore level 6 exception registers from a and r7

	; Now safe to use RETE
	RETE

\end{lstlisting}

However, if the stack pointer is invalid in this case, the double fault will still cause the CPU state to be corrupted.
You could provide an even more robust implementation by storing/loading from hardcoded memory addresses that are known
to always be valid.

\subsubsection{Fault Metadata Register Format}
\label{sec:fault-metadata-register-format}

The fault metadata register (register 7) stores additional information about faults beyond just the return address. The
return\_status\_register field is used to encode fault-specific metadata in the following format:

\begin{itemize}
	\item \textbf{Bits 0-2:} Bus access type (\texttt{BAT0}--\texttt{BAT2}) captured when the fault was raised (3 bits)
	      \begin{itemize}
		      \item 0x0: None
		      \item 0x1: Instruction fetch
		      \item 0x2: Data read
		      \item 0x3: Data write
		      \item 0x4: Exception vector fetch
		      \item 0x5: DMA read burst
		      \item 0x6: DMA write burst
		      \item 0x7: Reserved
	      \end{itemize}
	\item \textbf{Bit 3:} Double fault flag (1 bit) - set to 1 if this is a double fault
	\item \textbf{Bits 4-7:} Current fault type (4 bits) - the type of fault that occurred
	\item \textbf{Bits 8-11:} Original fault type (4 bits) - only meaningful for double faults, indicates the fault type
	      that was being handled when the double fault occurred
	\item \textbf{Bits 12-15:} Unused (reserved for future use)
\end{itemize}

Fault types use the following values:

\begin{itemize}
	\item \textbf{0x01:} Bus fault
	\item \textbf{0x02:} Alignment fault
	\item \textbf{0x03:} Segment overflow fault
	\item \textbf{0x04:} Invalid opcode fault
	\item \textbf{0x05:} Privilege violation fault
	\item \textbf{0x06:} Instruction trace fault
	\item \textbf{0x07:} Level five hardware exception conflict
	\item \textbf{0x08:} Double fault
	\item \textbf{0x09:} Bus protection fault
\end{itemize}

The return\_address field of register 7 stores the bus address that was being accessed when the fault occurred (for
faults that involve memory access). For coprocessor invalid-opcode faults that occur before bus access, this field will
be zero.

Exception handlers can use the \texttt{ETFR} instruction to read the fault metadata register and decode this
information to determine how to handle the fault appropriately.

\subsection{Cause Register}

The cause register is a 16-bit internal register used to communicate between different CPU units and coordinate
exception handling. It has the following structure:

\begin{itemize}
	\item \textbf{Bits 15-12 (First Nibble):} Coprocessor ID (0x1 for exception unit, 0x0 for processing unit)
	\item \textbf{Bits 11-8 (Second Nibble):} Coprocessor opcode (the operation to execute)
	\item \textbf{Bits 7-0 (Third and Fourth Nibbles):} Vector or parameter value
\end{itemize}

The cause register is automatically populated by the exception unit when a pending exception needs to be serviced. The
priority of exceptions is determined by examining the first nibble of the vector ID, calculated as 7 minus the value of
the first nibble (e.g., 0x00 is priority 7, 0x40 is priority 3, 0x60 and above are all priority 1).

\subsection{Hardware Interrupt Enable Bits}

The status register contains 4 individual enable bits (bits 9-12 in the privileged byte) that control whether each
maskable hardware interrupt line can trigger an exception:

\begin{itemize}
	\item Bit 9 (E1): Hardware Interrupt Line 1 (Level 1, lowest maskable priority)
	\item Bit 10 (E2): Hardware Interrupt Line 2 (Level 2)
	\item Bit 11 (E3): Hardware Interrupt Line 3 (Level 3)
	\item Bit 12 (E4): Hardware Interrupt Line 4 (Level 4, highest maskable priority)
\end{itemize}

\textbf{Note:} Hardware Interrupt Line 5 (Level 5) is a Non-Maskable Interrupt (NMI) and cannot be disabled.
It will always trigger an exception when asserted, regardless of the enable bit settings. Bit 13 is used for the
Exception Active flag instead.

When an interrupt line is disabled (bit = 0), interrupts on that line are completely ignored and not queued. When
enabled (bit = 1), the interrupt can trigger an exception if its priority is higher than the currently executing
exception level.

\subsection{Pending Hardware Interrupts}

Hardware interrupt pins are level-sensitive inputs. When an enabled interrupt line is observed, the corresponding bit
is set in the pending hardware-exceptions register. Disabled maskable interrupt lines are ignored at this sampling
point and are not latched for later service.

The pending register stores one bit per hardware interrupt line, so repeated assertions of the same line coalesce into
a single pending interrupt. If the external pin remains asserted after the interrupt is serviced, it may be sampled
again and become pending again.

At an instruction boundary, the exception unit selects the highest-priority pending hardware interrupt whose priority
is higher than the current exception level. The selected bit is cleared as the handler is dispatched. Pending
lower-priority interrupts remain latched and are considered again after the current handler returns with
\mnemonic{RETE}.

Faults have priority over all hardware interrupts. If a fault is pending at the same instruction boundary as one or
more hardware interrupts, the fault is dispatched first and the hardware interrupts remain pending. This includes
instruction trace faults: when trace mode raises a post-instruction fault, that trace fault is serviced before pending
hardware interrupts at the next boundary.

\textbf{Hierarchical Exception Handling:}

Exceptions in SIRC-1 are strictly hierarchical. A hardware interrupt can only interrupt the current execution if its
priority level (exception level) is \textit{higher} than the currently active exception level. This prevents:

\begin{itemize}
	\item An interrupt handler from being interrupted by its own interrupt (same level)
	\item Lower priority interrupts from interrupting higher priority handlers
	\item Software exceptions from running during any hardware interrupt (would corrupt link register 0)
\end{itemize}

The exception unit maintains a \texttt{current\_exception\_level} field that tracks the active exception level (0-7).
When not handling any exception, this is 0. When handling an exception at level N, only exceptions at level > N can
interrupt.

\textbf{Level 5 NMI-like Behavior:}

Level 5 hardware exceptions behave similarly to a Non-Maskable Interrupt (NMI). They cannot be disabled via the
interrupt enable bits in the status register, and will preempt any lower-priority exception handler. However, Level 5
is \textit{not} fully non-maskable. Because faults carry priority 7 (higher than Level 5's priority 6), a L5 interrupt
that arrives while a fault is being handled is \textit{queued} in the pending hardware exceptions register rather than
serviced immediately.

Once the fault handler returns via \texttt{RETE} and the exception level drops back below 6, the pending L5 interrupt
fires normally.

This means Level 5 response latency is not deterministic when a fault is in progress. For timing-critical use of the
NMI line, fault handlers should be kept as short as possible.

Keep in mind that if a second L5 interrupt arrives \textit{while the CPU is already executing an L5 handler} (exception
level 6), it is not queued. It will raise a Level Five Hardware Exception Conflict fault instead. See the fault
descriptions above.

These two rules compose in a non-obvious way: if a fault is raised \textit{inside} an L5 handler and a second L5
interrupt arrives while that fault is being serviced, the interrupt is queued (priority 7 blocks priority 6). Once the
fault handler returns via \texttt{RETE} and the exception level drops back to 6, the queued L5 fires — and since the
CPU is already at level 6 at that point, a Level Five Hardware Exception Conflict fault is raised then, not during the
inner fault handler.

\section{Exception Processing}

\subsection{Exception Priority}

Exceptions are prioritized as follows (highest to lowest):

\begin{enumerate}
	\item Priority 7: Faults (abort exceptions)
	\item Priority 6: Level 5 hardware exceptions (NMI-like)
	\item Priority 5: Level 4 hardware exceptions
	\item Priority 4: Level 3 hardware exceptions
	\item Priority 3: Level 2 hardware exceptions
	\item Priority 2: Level 1 hardware exceptions
	\item Priority 1: Software exceptions (traps)
\end{enumerate}

\subsection{Exception Handling Flow}

When an exception occurs, the following steps are executed:

\begin{enumerate}
	\item For hardware exceptions, the exception unit checks if the interrupt line is enabled in the status register. If
	      disabled, the interrupt is ignored and not queued.

	\item The exception unit checks if the exception priority level is higher than the current active exception level. If not,
	      the exception is deferred until the current handler completes.

	\item For level 5 hardware exceptions being raised when one is already being handled (both have exception level 6), a Level
	      Five Hardware Exception Conflict fault is raised instead.

	\item The current program counter, status register, and current exception level are stored in the appropriate link register
	      based on the exception priority level.

	\item The Protected Mode bit in the status register is cleared, switching the CPU to supervisor mode.

	\item The current exception level is updated to the level of the exception being serviced.

	\item The vector-table address is computed by multiplying the vector ID by 2 (since vectors are 32-bit addresses stored as
	      two 16-bit words), then adding the system RAM offset.

	\item The exception unit reads the high word and then the low word of the vector target address from the vector table.

	\item The program counter is set to the fetched vector target address, transferring control to the exception handler.
\end{enumerate}

\subsection{Exception Entry Side Effects}

For pending hardware interrupts, priority and mask checks occur before the exception-vector fetch. A hardware interrupt
that is disabled or not high enough priority to preempt the current handler does not fetch its vector; if it is already
latched as pending, it remains pending for later service according to the pending-interrupt rules above.

After an exception source is selected for service, the exception unit fetches the high word and low word of the handler
target address from the vector table. Architectural entry side effects commit only after the vector target has been
fetched and the exception is accepted for service.

When entry commits, the exception unit performs the following state updates as one architectural action:

\begin{itemize}
	\item Writes the selected link register with the current program counter, current status register, and current exception
	      level.
	\item Clears \texttt{SR.P}, entering supervisor mode.
	\item Clears \texttt{SR.T}, preventing trace faults from firing inside the handler.
	\item Sets \texttt{SR.EA}, indicating that an exception handler is active.
	\item Preserves the condition flags, reserved lower status bits, interrupt-enable bits, and \texttt{SR.A} in the live status
	      register.
	\item Updates the current exception level to the level of the exception being serviced.
	\item Loads the program counter from the fetched vector target address.
\end{itemize}

The saved status register in the link register is the pre-entry value, before \texttt{SR.P}, \texttt{SR.T}, or
\texttt{SR.EA} are modified. \mnemonic{RETE} restores that saved status register, so a handler can inspect or edit the
saved value with \mnemonic{ETFR} and \mnemonic{ETTR} before returning.

If an exception is not accepted because its priority is not higher than the current exception level, none of these
entry side effects occur. Pending hardware interrupts remain subject to the priority and pending-interrupt rules
described above.

The behavior of bus or protection faults during an exception-vector fetch is currently implementation-defined and
should not be relied upon by software.

\textbf{Returning from Exceptions:}

The RETE (Return from Exception) instruction restores the saved state from the link register:

\begin{enumerate}
	\item The return address is loaded from the link register into the program counter
	\item The saved status register value is restored (including interrupt enable bits)
	\item The saved exception level is restored to current\_exception\_level
	\item Execution resumes at the return address
\end{enumerate}

After RETE, any pending interrupts that are now eligible to run (higher priority than the restored exception level and
enabled) will be serviced before the next instruction executes.

\subsection{Vector Table}

The exception vectors are stored starting at address 0x0. Each vector is a 24-bit target address stored as two
consecutive 16-bit words in the address format described in Section~\ref{sec:exception-vector-storage}. The high word
is stored first at \texttt{vector * 2}, followed by the low word at \texttt{vector * 2 + 1}. The vector table layout
is:

\begin{itemize}
	\item \textbf{0x00:} Reset vector
	\item \textbf{0x01:} Bus fault
	\item \textbf{0x02:} Alignment fault
	\item \textbf{0x03:} Segment overflow fault
	\item \textbf{0x04:} Invalid opcode fault
	\item \textbf{0x05:} Privilege violation fault
	\item \textbf{0x06:} Instruction trace fault
	\item \textbf{0x07:} Level five hardware exception conflict
	\item \textbf{0x08:} Double fault
	\item \textbf{0x09:} Bus protection fault
	\item \textbf{0x0A-0x0F:} Reserved for future privileged exceptions
	\item \textbf{0x10:} Level 5 hardware exception vector
	\item \textbf{0x20:} Level 4 hardware exception vector
	\item \textbf{0x30:} Level 3 hardware exception vector
	\item \textbf{0x40:} Level 2 hardware exception vector
	\item \textbf{0x50:} Level 1 hardware exception vector
	\item \textbf{0x60-0xFF:} User exception vectors (128 user-accessible trap vectors)
\end{itemize}

\section{Reset and Startup}

Reset can be initiated externally by asserting the \texttt{RSTI} pin, or in software by executing the privileged
\mnemonic{RSET} meta-instruction. Both paths converge on the same reset-vector sequence: the CPU holds reset output
active, then the exception unit fetches vector 0x00 and loads the fetched target address into the program counter.

\subsection{Reset State}

\begin{table}[H]
	\centering
	\begin{tabular}{lp{0.62\textwidth}}
		\toprule
		\textbf{State}                         & \textbf{Reset Behavior}                                                                              \\
		\midrule
		Instruction phase                      & Set to phase 0.                                                                                      \\
		Pending bus request                    & Cleared; an in-progress stalled bus request is abandoned.                                            \\
		Halt state                             & Cleared.                                                                                             \\
		Wait-for-exception state               & Cleared.                                                                                             \\
		Status register                        & Cleared to 0x0000 when the reset operation is executed by the exception unit.                        \\
		Program counter                        & Loaded from reset vector 0x00.                                                                       \\
		General registers                      & Not architecturally cleared by reset. Software should treat their contents as undefined after reset. \\
		Address register pairs                 & Not architecturally cleared by reset.                                                                \\
		Exception link registers               & Not architecturally cleared by reset.                                                                \\
		Pending hardware exceptions and faults & Cleared. External interrupt pins that remain asserted after reset may be
		sampled again according to the normal interrupt rules.                                                                                        \\
		Pending coprocessor command            & Set internally to the reset operation until the reset vector sequence completes.                     \\
		\bottomrule
	\end{tabular}
	\caption{Reset State}
	\label{tab:reset-state}
\end{table}

\subsection{Reset Output Hold}

While reset is active, the CPU asserts \texttt{RSTO} and performs no bus cycles.

For hardware reset, asserting \texttt{RSTI} immediately aborts any current instruction, clears any pending bus wait,
and asserts \texttt{RSTO}. If \texttt{RSTI} remains asserted, \texttt{RSTO} remains asserted and the CPU does not
advance. After \texttt{RSTI} is released, the reset hold completes before the CPU resumes. The reset assertion cycle
counts as the first reset-output cycle; after release, five additional countdown cycles occur before the exception unit
starts the reset-vector fetch.

For software reset, \mnemonic{RSET} completes as a normal coprocessor meta-instruction. When its write-back phase
commits the reset command, the CPU requests reset and \texttt{RSTO} is asserted for six cycles before the reset-vector
fetch begins.

\subsection{Reset Vector Fetch}

The reset vector is vector 0x00 in the exception vector table. After the reset-output hold, the exception unit performs
an exception-vector fetch with bus access type \texttt{ExceptionVectorFetch}:

\begin{enumerate}
	\item Read the high word of the reset target address from \texttt{system\_ram\_offset + 0x0000}.
	\item Read the low word of the reset target address from \texttt{system\_ram\_offset + 0x0001}.
	\item Clear the status register to 0x0000.
	\item Load the fetched target address into \reg{p}.
\end{enumerate}

The behavior of bus or protection faults during the reset-vector fetch is currently implementation-defined and should
not be relied upon by software.

\section{Return from Exception}

To return from an exception handler, the \texttt{RETE} (Return from Exception) instruction is used. This instruction
performs the following operations:

\begin{enumerate}
	\item Reads the link register corresponding to the current interrupt mask level (minus 1, since the mask is set to the
	      exception level)
	\item Restores the status register from the link register (including the interrupt mask and protected mode bit)
	\item Restores the program counter from the link register
	\item Execution continues from the restored address
\end{enumerate}

For abort exceptions (faults), \texttt{RETE} returns to the faulting instruction, allowing it to be retried. For
regular exceptions, \texttt{RETE} returns to the instruction following the one that was executing when the exception
occurred.

\subsection{Accessing Link Registers}

The exception coprocessor provides instructions to transfer data to and from the link registers:

\begin{itemize}
	\item \textbf{ETFR (Exception Transfer From Register):} Copies the return address and/or status register from a link register to the address register and/or R7
	\item \textbf{ETTR (Exception Transfer To Register):} Copies the address register and/or R7 to a link register's return address and/or status register
\end{itemize}

These instructions allow exception handlers to examine or modify the saved processor state before returning, enabling
features like system calls, context switching, and debuggers.

\section{Exception Coprocessor Instructions}

The exception coprocessor (coprocessor ID 0x1) provides several specialized instructions for exception handling and
system control.

\begin{table}[H]
	\centering
	\begin{tabular}{llll}
		\toprule
		\textbf{Instruction} & \textbf{Opcode} & \textbf{Privilege} & \textbf{Purpose}                                 \\
		\midrule
		\mnemonic{EXCP}      & 0x1             & User               & Trigger a software exception (trap)              \\
		\mnemonic{WAIT}      & 0x9             & Supervisor         & Enter wait state until an exception occurs       \\
		\mnemonic{RETE}      & 0xA             & Supervisor         & Return from an exception handler                 \\
		\mnemonic{RSET}      & 0xB             & Supervisor         & Perform reset sequence and jump to reset vector  \\
		\mnemonic{ETFR}      & 0xC             & Supervisor         & Copy saved exception state from a link register  \\
		\mnemonic{ETTR}      & 0xD             & Supervisor         & Write CPU register state back to a link register \\
		\bottomrule
	\end{tabular}
	\caption{Exception Coprocessor Instruction Summary}
	\label{tab:exception-coprocessor-summary}
\end{table}

For full syntax, operation details, and worked examples, see Chapter~\ref{ch:meta-instructions}.

\subsection{Internal Instructions}

The following opcodes are used internally by the exception unit and are not directly accessible via user instructions:

\begin{itemize}
	\item \textbf{Fault (0xE):} Automatically invoked when a fault condition is detected
	\item \textbf{HardwareException (0xF):} Automatically invoked when a hardware interrupt is signaled
	\item \textbf{None (0x0):} No operation, used when no exception is pending
\end{itemize}

These internal operations follow the standard exception handling flow described in the Exception Processing section.
