\chapter{Addressing Modes}
\label{ch:addressing-modes}

\section{Overview}

The SIRC-1 CPU supports seven addressing modes that specify how operands are accessed. These modes determine whether an
operand is an immediate value, register contents, or data in memory at a computed address.

\begin{table}[H]
	\centering
	\begin{tabular}{lp{8cm}}
		\toprule
		\textbf{Mode}      & \textbf{Description}                                   \\
		\midrule
		Immediate          & Operand is a constant value encoded in the instruction \\
		Register Direct    & Operand is the contents of a specified register        \\
		Indirect Immediate & Memory address = address register + immediate offset   \\
		Indirect Register  & Memory address = address register + register offset    \\
		Post-Increment     & Indirect register, then increment address register     \\
		Pre-Decrement      & Decrement address register, then use as indirect       \\
		Short Immediate    & 8-bit constant with optional shift                     \\
		\bottomrule
	\end{tabular}
	\caption{Addressing Modes Summary}
	\label{tab:addr-modes}
\end{table}

\section{Addressing Mode Matrix}

\begin{table}[H]
	\centering
	\scriptsize
	\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{1.8cm}>{\raggedright\arraybackslash}X>{\raggedright\arraybackslash}X>{\raggedright\arraybackslash}p{1.9cm}>{\raggedright\arraybackslash}X}
		\toprule
		\textbf{Mode}      & \textbf{Syntax}                                  & \textbf{Effective value or address}                                                    & \textbf{Memory access}  & \textbf{Legal families}                              \\
		\midrule
		Immediate          & \texttt{\#imm16}                                 & 16-bit encoded value                                                                   & None                    & ALU, LOAD-immediate, COPI, branch meta-instructions  \\
		Short Immediate    & \texttt{\#imm8, shift}                           & 8-bit zero-extended value; shift applies to register source                            & None                    & ALU short-immediate, COPI short-immediate            \\
		Register Direct    & \texttt{rN}                                      & 16-bit register value                                                                  & None                    & ALU, LOAD register-copy, COPR, register-offset forms \\
		Indirect Immediate & \texttt{(\#offset, addr)}                        & \texttt{addr.high : (addr.low + offset)}                                               & LOAD reads; STOR writes & Memory, LDEA, LDEL                                   \\
		Indirect Register  & \texttt{(rO, addr)}                              & \texttt{addr.high : (addr.low + rO)}                                                   & LOAD reads; STOR writes & Memory, LDEA, LDEL                                   \\
		Post-Increment     & \texttt{(\#offset, addr)+}, \texttt{(rO, addr)+} & Address is calculated from original \texttt{addr}; \texttt{addr.low += 1} after access & LOAD reads              & LOAD, LDEL                                           \\
		Pre-Decrement      & \texttt{-(\#offset, addr)}, \texttt{-(rO, addr)} & \texttt{addr.low -= 1}; address then uses decremented \texttt{addr.low + offset}       & STOR writes             & STOR, LDEA                                           \\
		\bottomrule
	\end{tabularx}
	\caption{Addressing Mode Side Effects and Legal Instruction Families}
	\label{tab:addressing-mode-matrix}
\end{table}

\section{Operand Categories}

Instruction descriptions use these operand categories:

\begin{description}
	\item[Register value] A 16-bit general-purpose, status, or address-component register operand used directly by an ALU, load,
	      store, or coprocessor instruction.
	\item[Immediate value] A constant encoded in the instruction. Full immediates are 16 bits; short immediates are 8 bits.
	\item[Memory source] An effective address read by \mnemonic{LOAD}.
	\item[Memory destination] An effective address written by \mnemonic{STOR}.
	\item[Branch target] An effective address written to the program-counter pair \reg{p}.
	\item[Address destination] An address-register pair written by \mnemonic{LDEA} or \mnemonic{LDEL}.
	\item[Alterable destination] Any destination register, address-register pair, or memory location that the instruction can
	      modify when its condition is true.
\end{description}

\section{Common Address Calculation Rules}

For indirect, post-increment, pre-decrement, branch, jump, and effective-address forms, address arithmetic is performed
on the low word of the selected address-register pair. The high word supplies the segment and is preserved unless the
instruction explicitly writes an address-register pair destination.

Post-increment and pre-decrement change the selected address register by exactly one word. The displacement or index
register participates only in the effective address calculation; it does not change the auto-update amount.

When \reg{p} is used as the source address register, the value used for address calculation is the program counter
after the current instruction has been fetched. Since every instruction is two words, PC-relative branch displacements
are relative to the next instruction address, not to the address of the branch instruction itself.

\section{Immediate Addressing}

\subsection{Description}

The operand is a constant value 16-bit value encoded directly in the instruction.

\subsection{Syntax}

\begin{lstlisting}
ADDI r1, #100         ; 16-bit immediate
ADDI r2, #-50         ; Signed 16-bit immediate
\end{lstlisting}

\subsection{Effective Value}

16-bit value encoded into the instruction. Arithmetic instructions interpret this field according to
Chapter~\ref{ch:data-representation}.

\subsection{Usage}

\begin{itemize}
	\item Loading constant values
	\item Arithmetic with known values
	\item Bit manipulation with masks
\end{itemize}

\section{Register Direct Addressing}

\subsection{Description}

The operand is the contents of a specified register. This is the most common mode for ALU operations in a load/store
architecture.

\subsection{Syntax}

\begin{lstlisting}
ADDR r1, r2, r3   ; r1 = r2 + r3
ADDR r4, r5       ; r4 = r4 + r5
\end{lstlisting}

\subsection{Effective Value}

The 16-bit value currently stored in the specified register.

\subsection{Usage}

\begin{itemize}
	\item All ALU operations between registers
	\item Register-to-register data movement
	\item Testing register values
\end{itemize}

\section{Indirect Immediate Addressing}

\subsection{Description}

The effective address is computed by adding a 16-bit signed immediate offset to an address register pair low word.
Memory instructions use the computed address for a load or store; effective-address instructions write the computed
address to an address-register pair.

\subsection{Syntax}

\begin{lstlisting}
LOAD r1, (#8, a)      ; Load from address (a + 8)
STOR (#-4, s), r2     ; Store to address (s - 4)
LDEA p, (#100, p)     ; Jump relative (pc = pc + 100)
\end{lstlisting}

\subsection{Effective Address}

\begin{equation}
	\text{EA} = \text{AddressRegister.high} : (\text{AddressRegister.low} + \text{Offset})
\end{equation}

The address register can be: \reg{l} (link), \reg{a} (address), \reg{s} (stack), or \reg{p} (program counter).

\subsection{Usage}

\begin{itemize}
	\item Accessing structure fields (fixed offset from base)
	\item Array element access (if index is constant)
	\item Stack frame access (local variables)
	\item Position-independent code (PC-relative addressing)
\end{itemize}

\section{Indirect Register Addressing}

\subsection{Description}

Similar to indirect immediate, but the offset is taken from a register instead of being encoded as an immediate value.
This enables runtime-computed offsets.

\subsection{Syntax}

\begin{lstlisting}
LOAD r1, (r2, a)      ; Load from address (a + r2)
STOR (r4, s), r3      ; Store to address (s + r4)
\end{lstlisting}

\subsection{Effective Address}

\begin{equation}
	\text{EA} = \text{AddressRegister.high} : (\text{AddressRegister.low} + \text{IndexRegister})
\end{equation}

\subsection{Usage}

\begin{itemize}
	\item Array indexing with variable index
	\item Table lookups
	\item Dynamic offset calculations
\end{itemize}

\section{Post-Increment Addressing}

\subsection{Description}

The effective address is computed using indirect register addressing, then the address register pair is incremented
after the memory operation.

This mode is particularly useful for iterating through arrays or implementing stack operations.

\subsection{Syntax}

\begin{lstlisting}
LOAD r1, (r2, s)+     ; Load from (s + r2), then s = s + 1
LOAD r1, (#2, s)+     ; Load from (s + 2), then s = s + 1
\end{lstlisting}

\subsection{Operation Sequence}

\begin{enumerate}
	\item Compute EA = AddressRegister.high : (AddressRegister.low + Offset)
	\item Perform memory operation (load)
	\item AddressRegister.low = AddressRegister.low + 1
\end{enumerate}

\subsection{Usage}

\begin{itemize}
	\item Stack pop operations
	\item Forward array traversal
	\item Buffer reading
\end{itemize}

\section{Pre-Decrement Addressing}

\subsection{Description}

The address register pair is decremented before computing the effective address for the memory operation.

This mode is the complement of post-increment and is useful for stack push operations.

\subsection{Syntax}

\begin{lstlisting}
STOR -(r2, s), r1     ; s = s - 1, then store to (s + r2)
STOR -(#2, s), r3     ; s = s - 1, then store to (s + 2)
\end{lstlisting}

\subsection{Operation Sequence}

\begin{enumerate}
	\item AddressRegister.low = AddressRegister.low - 1
	\item Compute EA = AddressRegister.high : (AddressRegister.low + Offset)
	\item Perform memory operation (store)
\end{enumerate}

\subsection{Usage}

\begin{itemize}
	\item Stack push operations
	\item Backward array traversal
	\item Buffer writing
\end{itemize}

\section{Short Immediate Addressing}

\subsection{Description}

Similar to immediate addressing, but limited to 8-bit values so shift information can fit into the instruction
encoding.

\subsection{Syntax}

\begin{lstlisting}
ADDI r1, #10, LSL #2  ; r1 = (r1 << 2) + 10
SUBI r2, #1, LSL #8   ; r2 = (r2 << 8) - 1
\end{lstlisting}

\subsection{Effective Value}

8-bit value encoded into the instruction and zero-extended before ALU arithmetic.

The shift is applied to the source operand, not the constant.

\subsection{Usage}

\begin{itemize}
	\item Loading constant values
	\item Arithmetic with known values
	\item Bit manipulation with masks
\end{itemize}

\section{Address Register Selection}

Many addressing modes require specifying which address register pair to use. The 2-bit address register field encodes:

\begin{table}[H]
	\centering
	\begin{tabular}{ccl}
		\toprule
		\textbf{Bits} & \textbf{Symbol} & \textbf{Register Pair}    \\
		\midrule
		00            & l               & Link Register (lh, ll)    \\
		01            & a               & Address Register (ah, al) \\
		10            & s               & Stack Pointer (sh, sl)    \\
		11            & p               & Program Counter (ph, pl)  \\
		\bottomrule
	\end{tabular}
	\caption{Address Register Encoding}
	\label{tab:addr-reg-encoding}
\end{table}

\section{Addressing Mode Examples}

\subsection{Structure Access}

\begin{lstlisting}
; Assume 'a' points to a struct:
; struct Point {
;   word 0: int16_t x
;   word 1: int16_t y
; }

LOAD r1, (#0, a)      ; r1 = point.x
LOAD r2, (#1, a)      ; r2 = point.y
\end{lstlisting}

\subsection{Array Iteration}

\begin{lstlisting}
; Iterate array of 16-bit values
; Assume 'a' points to array start
; r7 = counter
loop:
  LOAD r1, (#0, a)+   ; Load element, advance by 1
  ; ... process r1 ...
  SUBI r7, #1
  BRAN|!= @loop
\end{lstlisting}

\subsection{Stack Frame}

\begin{lstlisting}
:begin_subroutine

; Function prologue - save registers
STOR -(#0, s), r1
STOR -(#0, s), r2
STOR -(#0, s), r3

; Access parameters (above saved regs)
LOAD r4, (#3, s)      ; First parameter
LOAD r5, (#4, s)      ; Second parameter

; Function epilogue - restore registers
LOAD r3, (#0, s)+
LOAD r2, (#0, s)+
LOAD r1, (#0, s)+
RETS
\end{lstlisting}

\section{Addressing Mode Restrictions}

\subsection{Privilege Mode Restrictions}

Protected mode restricts writes to the high word of an address register:

\begin{itemize}
	\item \textbf{Direct high-register writes}: Trigger a privilege violation fault
	\item \textbf{Address-register-pair write-back}: Allowed only when the high word would remain unchanged; otherwise triggers a privilege violation fault
	\item \textbf{Post-increment/Pre-decrement}: Allowed in protected mode only when the high word would remain unchanged
\end{itemize}

\subsection{Alignment}

\begin{itemize}
	\item All memory addresses refer to 16-bit words
	\item There are no alignment restrictions for general memory I/O
	\item All instructions span 2 words and must begin at an even word address
	\item Unaligned instruction fetches (odd word address) trigger an alignment fault exception
\end{itemize}
