\chapter{Reading Instruction Descriptions}
\label{ch:reading-instructions}

This part of the manual describes each SIRCIS instruction in a consistent format. The goal is to make each instruction
entry usable as an implementation contract, not only as programmer guidance.

\section{Instruction Entry Fields}

\begin{description}
	\item[Mnemonic] The assembly mnemonic or mnemonic family. Immediate and register variants are listed together when they
	      perform the same operation.
	\item[Opcodes] The 6-bit opcode values for each supported instruction format.
	\item[Syntax] The accepted assembler forms. Meta-instructions are called out separately from real CPU instruction encodings.
	\item[Operands] The permitted operand kinds and addressing modes for the instruction.
	\item[Operation] Pseudocode for the architectural effect of the instruction. Arithmetic is performed on 16-bit word values
	      unless otherwise stated.
	\item[Write-back] Whether the computed result is written to a register, discarded, or used to update control state.
	\item[Status Flags] The effect on the condition flags in the status register.
	\item[Condition Codes] All normal instructions may be conditional. If the condition is false, the instruction has no
	      architectural side effects unless a later instruction description explicitly states otherwise.
	\item[Exceptions] Faults or traps that can be raised by the instruction. When an exception can occur, the entry should state
	      whether partial side effects are possible.
	\item[Timing] The base execution time and any circumstances that add wait states.
	\item[Privilege] Whether the instruction is available in protected mode, supervisor mode, or both.
\end{description}

\section{Notation}

\begin{table}[H]
	\centering
	\begin{tabular}{ll}
		\toprule
		\textbf{Notation} & \textbf{Meaning}                                       \\
		\midrule
		\texttt{rD}       & Destination general-purpose register                   \\
		\texttt{rS}       & Source general-purpose register                        \\
		\texttt{rS1}      & First source register, usually the left-hand operand   \\
		\texttt{rS2}      & Second source register, usually the right-hand operand \\
		\texttt{\#imm16}  & 16-bit immediate word encoded in immediate format      \\
		\texttt{\#imm8}   & 8-bit immediate encoded in short immediate format      \\
		\texttt{shift}    & Optional SIRCIS shift definition                       \\
		\texttt{SR.X}     & Status register field \texttt{X}                       \\
		\bottomrule
	\end{tabular}
	\caption{Instruction Description Notation}
	\label{tab:instruction-description-notation}
\end{table}

\section{Status Flag Effect Symbols}
\label{sec:flag-effect-symbols}

Flag tables use the following symbols:

\begin{table}[H]
	\centering
	\begin{tabular}{cp{9cm}}
		\toprule
		\textbf{Symbol} & \textbf{Meaning}                                          \\
		\midrule
		\texttt{*}      & Updated from the instruction result                       \\
		\texttt{0}      & Cleared to zero                                           \\
		\texttt{1}      & Set to one                                                \\
		\texttt{-}      & Preserved                                                 \\
		\texttt{S}      & Updated from the shifter result when \texttt{[S]} is used \\
		\bottomrule
	\end{tabular}
	\caption{Status Flag Effect Symbols}
	\label{tab:flag-effect-symbols}
\end{table}

\section{Conditional Execution}

Every encoded instruction contains a condition-code field. If the condition evaluates true, the instruction executes
normally. If the condition evaluates false, the instruction is skipped: destination registers, address registers,
memory, the program counter, link register, coprocessor state, and status flags are not modified by that instruction.

\section{Status Update Overrides}

ALU instructions can select the source of status flag updates with the status update field:

\begin{table}[H]
	\centering
	\begin{tabular}{lll}
		\toprule
		\textbf{Syntax} & \textbf{Encoded AF} & \textbf{Effect}                      \\
		\midrule
		\texttt{[N]}    & \texttt{00}         & Preserve status flags                \\
		\texttt{[A]}    & \texttt{01}         & Update flags from the ALU result     \\
		\texttt{[S]}    & \texttt{10}         & Update flags from the shifter result \\
		Reserved        & \texttt{11}         & Reserved for future use              \\
		\bottomrule
	\end{tabular}
	\caption{Status Update Source Encoding}
	\label{tab:status-update-source-encoding}
\end{table}

When no override is written, ALU instructions use \texttt{[A]}. Instructions that do not support status updates must
not use status update override syntax.
