\chapter{Memory Access Instructions}
\label{ch:memory-instructions}

\section{Overview}

Memory access instructions transfer data between registers and memory. Following the load/store architecture, these are
the only instructions that can access memory—all arithmetic and logical operations must work on registers.

The SIRC-1 provides two primary memory operations:
\begin{itemize}
	\item \mnemonic{LOAD} -- Read from memory into a register
	\item \mnemonic{STOR} -- Write from a register to memory
\end{itemize}

Both instructions support four memory addressing forms:
\begin{itemize}
	\item Indirect with immediate offset: \texttt{(\#offset, addr)}
	\item Indirect with register offset: \texttt{(reg, addr)}
	\item Post-increment load: \texttt{(\#offset, addr)+} or \texttt{(reg, addr)+}
	\item Pre-decrement store: \texttt{-(\#offset, addr)} or \texttt{-(reg, addr)}
\end{itemize}

\textbf{Important:} Register-offset memory forms support optional shift operations on the data being loaded or stored.
Immediate-offset memory forms do not encode a shift field. See Section~\ref{sec:memory-shifts} for details. Shifts are
also \textbf{not} supported when loading immediate values or copying register values directly.

\section{Effective Address Calculation}

For all memory operations, the effective address (EA) is calculated as:

\begin{equation}
	\text{EA} = \text{AddressRegister} + \text{Offset}
\end{equation}

Where:
\begin{itemize}
	\item Address Register is one of: \reg{l}, \reg{a}, \reg{s}, or \reg{p}
	\item Offset is either an immediate value or register contents
\end{itemize}

The resulting 24-bit address is used to access memory.

\section{Legal Forms}

\begin{table}[H]
	\centering
	\small
	\begin{tabular}{llll}
		\toprule
		\textbf{Instruction} & \textbf{Syntax}                        & \textbf{Encoding} & \textbf{Notes}                                  \\
		\midrule
		LOAD                 & \texttt{LOAD rD, (\#offset, addr)}     & 0x14              & Immediate offset                                \\
		LOAD                 & \texttt{LOAD rD, (rO, addr)[, shift]}  & 0x15              & Register offset; optional load-result shift     \\
		LOAD                 & \texttt{LOAD rD, (\#offset, addr)+}    & 0x16              & Immediate offset; post-increment                \\
		LOAD                 & \texttt{LOAD rD, (rO, addr)+[, shift]} & 0x17              & Register offset; optional shift; post-increment \\
		STOR                 & \texttt{STOR (\#offset, addr), rS}     & 0x10              & Immediate offset                                \\
		STOR                 & \texttt{STOR (rO, addr), rS[, shift]}  & 0x11              & Register offset; optional source shift          \\
		STOR                 & \texttt{STOR -(\#offset, addr), rS}    & 0x12              & Immediate offset; pre-decrement                 \\
		STOR                 & \texttt{STOR -(rO, addr), rS[, shift]} & 0x13              & Register offset; optional shift; pre-decrement  \\
		\bottomrule
	\end{tabular}
	\caption{Legal Memory Instruction Forms}
	\label{tab:memory-legal-forms}
\end{table}

\noindent\textbf{Zero displacement:} Immediate-displacement memory forms may omit \texttt{\#0}. Thus
\texttt{LOAD rD, (addr)}, \texttt{LOAD rD, (addr)+}, \texttt{STOR (addr), rS}, and \texttt{STOR -(addr), rS} assemble
as the matching \texttt{\#0} forms. This shorthand does not add new addressing modes: \mnemonic{LOAD} has no
pre-decrement form and \mnemonic{STOR} has no post-increment form.

\section{Common Memory Semantics}

\begin{table}[H]
	\centering
	\begin{tabular}{lp{9cm}}
		\toprule
		\textbf{Topic}      & \textbf{Definition}                                                                                                                                                                                                                            \\
		\midrule
		Condition false     & The instruction is skipped and has no side effects. Registers, memory, address registers, and status flags are preserved.                                                                                                                      \\
		Address calculation & Immediate-offset forms add the 16-bit encoded offset to the selected address register low word. Register-offset forms add \texttt{rO} to the selected address register low word. The selected address register high word supplies the segment. \\
		Post-increment LOAD & The memory address is calculated from the original address register value. The address register low word is incremented by one word in write-back after the memory access.                                                                     \\
		Pre-decrement STOR  & The memory address uses the selected address register low word decremented by one word, plus the offset. The address register low word is decremented by one word in write-back after the memory access.                                       \\
		Status flags        & Memory instructions preserve all status flags. Shifts used by memory instructions cannot update the status register.                                                                                                                           \\
		Timing              & Memory instructions complete in 6 cycles when the bus access is acknowledged without wait states. Bus wait states extend the memory-access phase.                                                                                              \\
		Exceptions          & Memory instructions can raise data bus, data bus-protection, and segment-overflow faults; see the notes below.                                                                                                                                 \\
		Fault side effects  & Destination register writes and address-register auto-updates occur in write-back after the memory-access phase. If a fault aborts execution before write-back, those write-back effects do not occur.                                         \\
		\bottomrule
	\end{tabular}
	\caption{Common Memory Instruction Semantics}
	\label{tab:memory-common-semantics}
\end{table}

\noindent Memory instructions can raise data bus and data bus-protection faults during the memory-access phase.
Effective-address calculation can raise a segment-overflow fault when \texttt{SR.A} (Trap on Address Overflow) is set
and the 16-bit low-word address calculation wraps. Documented memory forms do not raise privilege-violation or invalid-opcode faults. Instruction
fetch can still raise the normal fetch faults before a memory instruction executes. If trace mode was enabled when the
instruction began, an instruction-trace fault is raised after the instruction commits. Processing-unit encodings outside
the documented memory forms are reserved or architecturally undefined; they are not required to raise an invalid-opcode
fault.

\section{LOAD Instructions}

\begin{instructionbox}{LOAD -- Load from Memory}
	\textbf{Opcodes:} 0x14 (Indirect Imm), 0x15 (Indirect Reg), 0x16 (Post-Inc Imm), 0x17 (Post-Inc Reg)

	\textbf{Syntax:}
	\begin{lstlisting}
LOAD rD, (#offset, addr)      ; rD = memory[addr + offset]
LOAD rD, (addr)               ; equivalent to LOAD rD, (#0, addr)
LOAD rD, (rS, addr)           ; rD = memory[addr + rS]
LOAD rD, (#offset, addr)+     ; rD = memory[addr + offset]; addr += 1
LOAD rD, (addr)+              ; equivalent to LOAD rD, (#0, addr)+
LOAD rD, (rS, addr)+          ; rD = memory[addr + rS]; addr += 1

; With optional shift (register-offset memory forms only)
LOAD rD, (rS, addr), shift           ; Load and shift
LOAD rD, (rS, addr)+, shift          ; Load, shift, then increment
\end{lstlisting}

	\textbf{Operands:} \texttt{rD} is the destination general-purpose register. \texttt{addr} is the address-register
	pair that supplies the segment and base low word. Immediate-offset forms use a 16-bit displacement. Register-offset
	forms use \texttt{rS} as the displacement register and may apply a shift to the loaded memory value before write-back.

	\textbf{Operation:}
	\begin{verbatim}
if post_increment then
    effective_address = addr + offset
    loaded_value = memory[effective_address]
    destination = shift_if_encoded(loaded_value)
    addr.low = addr.low + 1
else
    effective_address = addr + offset
    loaded_value = memory[effective_address]
    destination = shift_if_encoded(loaded_value)
endif
; Flags are preserved
\end{verbatim}

	\textbf{Description:}

	Loads a 16-bit value from memory into the destination register. Post-increment forms update the address register after
	the load, useful for array traversal.

	\textbf{Write-back:} If the condition is true, writes the loaded value to \texttt{rD}. Post-increment forms also write
	the incremented address-register low word.

	\textbf{Flags:} Preserved.

	\textbf{Exceptions:} May raise data bus or data bus-protection faults during the memory read. May raise a
	segment-overflow fault during effective-address calculation when \texttt{SR.A} is set.

	\textbf{Timing:} 6 cycles when the memory read is acknowledged without wait states; wait states extend the memory-access
	phase.

	\textbf{Condition codes:} If the condition is false, no memory read occurs, no destination register is written, no
	address register is incremented, and status flags are preserved.

	\textbf{Privilege:} Available in protected mode and supervisor mode. Protected-mode data-memory access is subject to
	bus protection.

	\textbf{Examples:}
	\begin{lstlisting}
; Load from fixed offset
LOAD r1, (#4, a)              ; r1 = memory[a + 4]

; Load using variable offset
LOAD r2, (r3, a)              ; r2 = memory[a + r3]

; Array iteration
LOAD r4, (#0, a)+             ; r4 = *a; a += 1

; Stack pop
LOAD r5, (#0, s)+             ; r5 = *s; s += 1

; Load and shift (for bit manipulation)
LOAD r3, (r2, a), LSL #2      ; r3 = memory[a + r2] << 2
LOAD r6, (r2, a), ASR #2      ; r6 = memory[a + r2] >> 2 (signed)
\end{lstlisting}

\end{instructionbox}

\section{STOR Instructions}

\begin{instructionbox}{STOR -- Store to Memory}
	\textbf{Opcodes:} 0x10 (Indirect Imm), 0x11 (Indirect Reg), 0x12 (Pre-Dec Imm), 0x13 (Pre-Dec Reg)

	\textbf{Syntax:}
	\begin{lstlisting}
STOR (#offset, addr), rS      ; memory[addr + offset] = rS
STOR (addr), rS               ; equivalent to STOR (#0, addr), rS
STOR (rD, addr), rS           ; memory[addr + rD] = rS
STOR -(#offset, addr), rS     ; addr -= 1; memory[addr + offset] = rS
STOR -(addr), rS              ; equivalent to STOR -(#0, addr), rS
STOR -(rD, addr), rS          ; addr -= 1; memory[addr + rD] = rS

; With optional shift (register-offset memory forms only)
STOR (rD, addr), rS, shift           ; Shift and store
STOR -(rD, addr), rS, shift          ; Decrement, shift, and store
\end{lstlisting}

	\textbf{Operands:} \texttt{rS} is the source general-purpose register. \texttt{addr} is the address-register pair that
	supplies the segment and base low word. Immediate-offset forms use a 16-bit displacement. Register-offset forms use
	\texttt{rD} as the displacement register and may apply a shift to the source value before storing it.

	\textbf{Operation:}
	\begin{verbatim}
if pre_decrement then
    decremented_low = addr.low - 1
    effective_address = (addr.high : decremented_low) + offset
    memory[effective_address] = shift_if_encoded(source)
    addr.low = decremented_low
else
    effective_address = addr + offset
    memory[effective_address] = shift_if_encoded(source)
endif
; Source register and flags are preserved
\end{verbatim}

	\textbf{Description:}

	Stores a 16-bit value from a register into memory. Pre-decrement forms use the decremented address for the store and
	write the decremented address register value back after the memory access, useful for stack push operations.

	\textbf{Write-back:} If the condition is true, writes the source value to memory. Pre-decrement forms also write the
	decremented address-register low word.

	\textbf{Flags:} Preserved.

	\textbf{Exceptions:} May raise data bus or data bus-protection faults during the memory write. May raise a
	segment-overflow fault during effective-address calculation when \texttt{SR.A} is set.

	\textbf{Timing:} 6 cycles when the memory write is acknowledged without wait states; wait states extend the memory-access
	phase.

	\textbf{Condition codes:} If the condition is false, no memory write occurs, no address register is decremented, and
	status flags are preserved.

	\textbf{Privilege:} Available in protected mode and supervisor mode. Protected-mode data-memory access is subject to
	bus protection.

	\textbf{Examples:}
	\begin{lstlisting}
; Store to fixed offset
STOR (#8, a), r1              ; memory[a + 8] = r1

; Store using variable offset
STOR (r3, a), r2              ; memory[a + r3] = r2

; Store with shift (data shifted before storing)
STOR (r2, a), r1, LSL #1      ; memory[a + r2] = r1 << 1
STOR (r3, a), r1, LSR #3      ; memory[a + r3] = r1 >> 3

; Stack push
STOR -(#0, s), r4             ; s -= 1; memory[s] = r4
\end{lstlisting}

\end{instructionbox}

\section{Common Memory Access Patterns}

\subsection{Structure Field Access}
\begin{lstlisting}
; struct { int16_t x, y, z; } point;
; 'a' points to point
LOAD r1, (#0, a)              ; r1 = point.x
LOAD r2, (#1, a)              ; r2 = point.y
LOAD r3, (#2, a)              ; r3 = point.z
\end{lstlisting}

\subsection{Array Iteration}
\begin{lstlisting}
; Process array of 16-bit values
; 'a' = array base, r7 = count
loop:
    LOAD r1, (#0, a)+         ; Load element, advance
    ; ... process r1 ...
    SUBI r7, #1
    BRAN|!= @loop
\end{lstlisting}

\subsection{Stack Operations}
\begin{lstlisting}
; Push registers
STOR -(#0, s), r1
STOR -(#0, s), r2
STOR -(#0, s), r3

; Pop registers (reverse order)
LOAD r3, (#0, s)+
LOAD r2, (#0, s)+
LOAD r1, (#0, s)+
\end{lstlisting}

\section{Shift Operations with Memory Instructions}
\label{sec:memory-shifts}

Register-offset memory forms of \mnemonic{LOAD} and \mnemonic{STOR} support optional shift operations. This feature
allows data to be shifted during the load or store operation, which is particularly useful for bit manipulation and
working with the CPU's word-addressed memory architecture.

\subsection{LOAD with Shift}

When loading from memory, the data read from memory is shifted \textbf{after} being loaded but \textbf{before} being
written to the destination register:

\begin{lstlisting}
LOAD r3, (r2, a), LSL #2      ; r3 = (memory[a + r2]) << 2
LOAD r5, (r2, a), ASL #3      ; r5 = (memory[a + r2]) << 3
LOAD r6, (r4, a), ASR #2      ; r6 = (memory[a + r4]) >> 2 (sign-extend)
\end{lstlisting}

\subsection{STOR with Shift}

When storing to memory, the source register value is shifted \textbf{before} being written to memory. The original
register value is \textbf{not} modified:

\begin{lstlisting}
STOR (r2, a), r1, LSL #1      ; memory[a + r2] = r1 << 1 (r1 unchanged)
STOR (r2, a), r1, ASL #2      ; memory[a + r2] = r1 << 2 (r1 unchanged)
STOR (r4, a), r1, ASR #4      ; memory[a + r4] = r1 >> 4 (r1 unchanged)
\end{lstlisting}

\subsection{Supported Shift Types}

All standard shift types are supported:
\begin{itemize}
	\item \textbf{LSL} -- Logical Shift Left (fill with zeros)
	\item \textbf{LSR} -- Logical Shift Right (fill with zeros)
	\item \textbf{ASL} -- Arithmetic Shift Left (same as LSL)
	\item \textbf{ASR} -- Arithmetic Shift Right (sign-extend)
	\item \textbf{RTL} -- Rotate Left within the operand
	\item \textbf{RTR} -- Rotate Right within the operand
\end{itemize}

\subsection{Shift Limitations}

\textbf{Important:} Shift operations are \textbf{only} supported when using register-offset memory forms. The following modes do \textbf{not} support shifts:
\begin{itemize}
	\item Loading immediate values: \texttt{LOAD r1, \#0x1234} -- No shift allowed
	\item Copying registers: \texttt{LOAD r1, r2} -- No shift allowed
	\item Immediate-offset memory load: \texttt{LOAD r1, (\#0, a)} -- No shift allowed
	\item Immediate-offset memory store: \texttt{STOR (\#0, a), r1} -- No shift allowed
\end{itemize}

For unsupported shift forms, load or store the unshifted value and use an ALU instruction or \mnemonic{SHFT} for the
shift operation.

\subsection{Use Cases}

\subsubsection{Packed Data}

Shifting during load/store can save cycles when extracting or placing fields in packed words:

\begin{lstlisting}
; Load a packed word and move the high byte into the low byte position
LOAD r3, (r2, a), LSR #8
\end{lstlisting}

\subsubsection{Bit Manipulation}

Shifting during memory access enables efficient bit manipulation:

\begin{lstlisting}
; Extract upper byte from memory
LOAD r1, (r3, a), LSR #8      ; r1 = (memory[a + r3] >> 8) & 0xFF

; Store scaled value
STOR (r3, a), r2, LSL #2      ; memory[a + r3] = r2 * 4
\end{lstlisting}

\subsubsection{Combined with Auto-Increment/Decrement}

Shifts work seamlessly with post-increment and pre-decrement modes:

\begin{lstlisting}
; Load, shift, and advance pointer
LOAD r3, (r2, a)+, LSL #1     ; r3 = (memory[a + r2]) << 1; a += 1

; Decrement, shift, and store
STOR -(r2, a), r1, LSR #2     ; a -= 1; memory[a + r2] = r1 >> 2
\end{lstlisting}

\subsection{Status Flags}

\textbf{Note:} Unlike shifts in ALU instructions, shifts performed during \mnemonic{LOAD} and \mnemonic{STOR} operations can \textbf{not} update the status register flags. If you need flag updates based on a shift result, use an ALU instruction or the \mnemonic{SHFT} meta-instruction (see Chapter~\ref{ch:meta-instructions}).
