\chapter{Coprocessor Instructions}
\label{ch:coprocessor}

\section{Overview}

The SIRC-1 CPU includes a coprocessor interface that allows delegation of certain operations to specialized
coprocessors. The primary coprocessor is the Exception Unit (Coprocessor 1), which handles exceptions, interrupts, and
privileged operations.

Coprocessor calls are ordinary conditional instructions. If the instruction's condition is true, the processing unit
computes a command operand and writes it to the 16-bit pending coprocessor-command latch during write-back. The
selected coprocessor then interprets the command value. If the selected coprocessor is not present in the CPU model or
does not implement the requested operation, the exception unit raises an invalid-opcode fault so the operation can be
emulated or rejected by software.

\section{Legal Forms}

\begin{table}[H]
	\centering
	\small
	\begin{tabular}{llll}
		\toprule
		\textbf{Instruction} & \textbf{Syntax}       & \textbf{Encoding} & \textbf{Notes}                           \\
		\midrule
		COPI                 & \texttt{COPI \#imm16} & 0x0F              & 16-bit coprocessor command operand       \\
		COPR                 & \texttt{COPR rS}      & 0x3F              & Command operand is read from \texttt{rS} \\
		\bottomrule
	\end{tabular}
	\caption{Legal Coprocessor Instruction Forms}
	\label{tab:coprocessor-legal-forms}
\end{table}

\noindent The CPU encoding reserves register and shift fields inherited from the immediate and
register instruction formats. The programmer-facing \mnemonic{COPI} and \mnemonic{COPR} syntax hides those fields;
assemblers emit canonical values and coprocessors must not interpret them as destination-register writes. Opcode 0x2F has
no public assembly syntax and is undocumented.

\section{Common Semantics}

\begin{table}[H]
	\centering
	\begin{tabular}{lp{9cm}}
		\toprule
		\textbf{Topic}     & \textbf{Definition}                                                                                                                                                                                                              \\
		\midrule
		Condition false    & The instruction is skipped and has no side effects. The pending coprocessor command, registers, memory, address registers, program counter, and status flags are preserved.                                                      \\
		Command operand    & \mnemonic{COPI} uses the 16-bit immediate value. \mnemonic{COPR} uses its source register value as the command operand.                                                                                                          \\
		Command format     & For standard coprocessor commands, bits 15--12 select the coprocessor ID, bits 11--8 select the coprocessor operation, and bits 7--0 are coprocessor-defined operand bits.                                                       \\
		Write-back         & The processing unit writes the command operand to the pending coprocessor-command latch. It does not write general-purpose registers, address registers, or data memory as part of the call. Hidden encoding fields are ignored. \\
		Status flags       & Preserved by the processing-unit coprocessor-call write-back path. A coprocessor operation may define its own later CPU-state effects; exception-unit effects are documented in Chapter~\ref{ch:exceptions}.                     \\
		Privilege          & In protected mode, coprocessor-operation nibbles 0x0--0x7 are user-callable and 0x8--0xF are supervisor-only. Privilege checks only apply when the instruction condition is true.                                                \\
		Timing             & The coprocessor-call instruction itself uses the normal 6 execution phases and performs no data-memory bus access. The selected coprocessor may perform follow-up work after the command is latched.                             \\
		Exceptions         & Supervisor-only operations can raise privilege-violation faults; missing coprocessors and unimplemented operations can raise invalid-opcode faults. See the notes below.                                                         \\
		Fault side effects & If a fault is raised while dispatching the command, software must not depend on the contents of the pending command latch after the fault.                                                                                       \\
		\bottomrule
	\end{tabular}
	\caption{Common Coprocessor Instruction Semantics}
	\label{tab:coprocessor-common-semantics}
\end{table}

\noindent Coprocessor-call instructions do not perform data-memory access and do not raise data bus, data
bus-protection, data alignment, or segment-overflow faults in documented forms. Instruction fetch can still raise the
normal fetch faults before the call executes. If trace mode was enabled when the instruction began, an instruction-trace
fault is raised after the processing-unit call commits, unless an earlier fault prevents the instruction from completing.
In protected mode, supervisor-only coprocessor operation nibbles are rejected before the command is dispatched. Missing
coprocessors and unimplemented coprocessor operations are detected after the command has been latched for dispatch and
raise invalid-opcode faults through the exception unit.

\begin{instructionbox}{COPI/COPR -- Coprocessor Call}
	\textbf{Opcodes:} 0x0F (Immediate), 0x3F (Register). Opcode 0x2F is undocumented.

	\textbf{Syntax:}
	\begin{lstlisting}
COPI #opcode                  ; Call coprocessor with 16-bit command
COPR rS                       ; Call coprocessor with command in rS
\end{lstlisting}

	\textbf{Operands:} \mnemonic{COPI} takes a 16-bit immediate command operand. \mnemonic{COPR} takes a general-purpose
	source register whose 16-bit value is used as the command operand. Hidden encoded register and shift fields are
	canonicalized by the assembler and are not programmer-visible operands.

	\textbf{Description:}

	Transfers a command to a coprocessor. For the standard command layout, the high nibble selects the coprocessor and the
	next nibble selects that coprocessor's operation. The remaining low byte is defined by the selected operation.

	\textbf{Operation:}
	\begin{lstlisting}
if condition_true then
    pending_coprocessor_command = command_operand
endif
\end{lstlisting}

	\textbf{Write-back:} Writes the command operand to the pending coprocessor-command latch.

	\textbf{Flags:} Preserved by the coprocessor-call instruction.

	\textbf{Exceptions:} May raise privilege-violation faults for supervisor-only operations in protected mode. May raise
	invalid-opcode faults for missing coprocessors and unimplemented coprocessor operations.

	\textbf{Condition codes:} If the condition is false, no command is dispatched, the pending command latch is preserved,
	no coprocessor privilege or opcode checks are performed, and status flags are preserved.

	\textbf{Timing:} 6 cycles, plus any instruction-fetch wait states.

	\textbf{Privilege:} Available in protected mode and supervisor mode. In protected mode, coprocessor-operation nibbles
	0x8--0xF are supervisor-only and raise privilege-violation faults when executed.

	\textbf{Examples:}
	\begin{lstlisting}
; Wait for interrupt (Exception Unit)
COPI #0x1900                  ; Coprocessor 1, operation 0x9
; Prefer the WAIT meta-instruction for normal assembly code.

; Return from exception
COPI #0x1A00                  ; Coprocessor 1, operation 0xA
; Prefer the RETE meta-instruction for normal assembly code.

; User exception (system call)
COPI #0x1180                  ; Coprocessor 1, operation 0x1, vector 0x80
; Prefer EXCP #0x80 for normal assembly code.
\end{lstlisting}

\end{instructionbox}

\section{Coprocessor Compatibility}

Standard coprocessor IDs are stable across SIRC-1 models. A CPU model may omit an optional coprocessor, but it must not
assign that coprocessor ID to an incompatible unit. If software calls an absent coprocessor or an operation that the
selected coprocessor does not implement, the CPU raises an invalid-opcode fault.

\begin{table}[H]
	\centering
	\small
	\begin{tabularx}{\textwidth}{llclX}
		\toprule
		\textbf{ID} & \textbf{Name}        & \textbf{Required} & \textbf{Operations} & \textbf{Compatibility rule}                          \\
		\midrule
		0x0         & Processing Unit      & Yes               & Internal            & Executes SIRCIS processing-unit instructions         \\
		0x1         & Exception Unit       & Yes               & 0x0--0xD            & Handles reset, faults, interrupts, traps, and links  \\
		0x2         & DMA Unit             & Yes               & 0x8--0xA            & Required standard DMA transfer unit                  \\
		0x3         & Integer Maths Unit   & No                & 0x0--0x3            & Optional standard integer multiply/divide unit       \\
		0x4--0xF    & Reserved/model space & No                & Model-defined       & Reserved unless a CPU model documents the assignment \\
		\bottomrule
	\end{tabularx}
	\caption{Standard Coprocessor IDs}
	\label{tab:standard-coprocessor-ids}
\end{table}

\subsection{Model and Revision Policy}

\begin{itemize}
	\item Required coprocessors and documented operations must remain backward compatible within the SIRC-1 family.
	\item Optional standard coprocessors keep their assigned IDs even on models that omit them.
	\item Reserved coprocessor IDs, reserved operation nibbles, and reserved operand values must raise invalid-opcode faults when
	      they are not implemented by the selected CPU model.
	\item Software may probe optional coprocessors by executing the documented operation and handling an invalid-opcode fault, or
	      by using a model-specific capability table if one is provided by system firmware.
	\item Processing-unit undocumented opcodes are a separate case: they are accepted by the processing unit and have
	      architecturally undefined effects unless a future CPU revision documents them.
\end{itemize}

\section{Exception Unit (Coprocessor 1)}

The Exception Unit handles:
\begin{itemize}
	\item Faults (abort exceptions, highest priority)
	\item Hardware exceptions (interrupts, multiple priority levels)
	\item Software exceptions (traps)
	\item Privileged operations
\end{itemize}

\subsection{Common Operations}

\begin{table}[H]
	\centering
	\begin{tabular}{llp{8cm}}
		\toprule
		\textbf{Opcode} & \textbf{Mnemonic} & \textbf{Operation}                        \\
		\midrule
		0x1100-0x11FF   & EXCP              & User exception (trap) at vector 0x60-0xFF \\
		0x1900          & WAIT              & Wait for interrupt                        \\
		0x1A00          & RETE              & Return from exception                     \\
		0x1B00          & RSET              & System reset                              \\
		0x1C00-0x1C7F   & ETFR              & Transfer from exception link register     \\
		0x1D00-0x1D7F   & ETTR              & Transfer to exception link register       \\
		\bottomrule
	\end{tabular}
	\caption{Exception Unit Operations}
\end{table}

\section{Exception Unit Meta-Instructions}

\begin{instructionbox}{EXCP -- User Exception Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x1100 | vector}

	\textbf{Syntax:} \texttt{EXCP \#vector}

	\textbf{Operands:} \texttt{\#vector} is the low 8-bit user exception vector selector.

	\textbf{Description:} Triggers a software exception at the specified user vector, normally 0x60--0xFF. Used by
	user-mode programs to invoke system calls or request supervisor-mode services.

	\textbf{Write-back:} Inherits \mnemonic{COPI} command dispatch to the exception unit.

	\textbf{Flags:} Preserved by the coprocessor-call instruction; exception entry effects are defined by the exception
	unit.

	\textbf{Exceptions:} Dispatches a software exception through the exception unit.

	\textbf{Condition codes:} If the condition is false, no trap is dispatched and status flags are preserved.

	\textbf{Timing:} Same as \mnemonic{COPI}: 6 cycles for the coprocessor call, plus exception-unit follow-up work.

	\textbf{Privilege:} User-callable in protected mode and supervisor mode.

	\textbf{Example:}
	\begin{lstlisting}
LOAD r1, #'A'                 ; Character to print
EXCP #0x80                    ; Trap to OS print routine
\end{lstlisting}
\end{instructionbox}

\begin{instructionbox}{WAIT -- Wait for Interrupt Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x1900}

	\textbf{Syntax:} \texttt{WAIT}

	\textbf{Operands:} None.

	\textbf{Description:} Puts the CPU into wait state until an exception occurs. Only valid in supervisor mode.

	\textbf{Write-back:} Inherits \mnemonic{COPI} command dispatch to the exception unit.

	\textbf{Flags:} Preserved by the coprocessor-call instruction.

	\textbf{Exceptions:} Raises a privilege-violation fault if executed in protected mode.

	\textbf{Condition codes:} If the condition is false, the CPU does not enter wait state and status flags are preserved.

	\textbf{Timing:} Same as \mnemonic{COPI}: 6 cycles before the wait-state request is dispatched.

	\textbf{Privilege:} Supervisor mode only.
\end{instructionbox}

\begin{instructionbox}{RETE -- Return from Exception Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x1A00}

	\textbf{Syntax:} \texttt{RETE}

	\textbf{Operands:} None.

	\textbf{Description:} Returns from an exception handler by restoring the saved program counter and status register.

	\textbf{Write-back:} Inherits \mnemonic{COPI} command dispatch to the exception unit.

	\textbf{Flags:} Restored by the exception unit as part of exception return.

	\textbf{Exceptions:} Raises a privilege-violation fault if executed in protected mode.

	\textbf{Condition codes:} If the condition is false, no exception return occurs and current status flags are
	preserved.

	\textbf{Timing:} Same as \mnemonic{COPI}: 6 cycles before exception-unit return processing.

	\textbf{Privilege:} Supervisor mode only.
\end{instructionbox}

\begin{instructionbox}{RSET -- System Reset Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x1B00}

	\textbf{Syntax:} \texttt{RSET}

	\textbf{Operands:} None.

	\textbf{Description:} Performs a software reset of the processor by clearing the status register and jumping to the
	reset vector.

	\textbf{Write-back:} Inherits \mnemonic{COPI} command dispatch to the exception unit.

	\textbf{Flags:} Reset by the exception unit as part of reset processing.

	\textbf{Exceptions:} Raises a privilege-violation fault if executed in protected mode.

	\textbf{Condition codes:} If the condition is false, no reset occurs and current status flags are preserved.

	\textbf{Timing:} Same as \mnemonic{COPI}: 6 cycles before reset processing begins.

	\textbf{Privilege:} Supervisor mode only.
\end{instructionbox}

\begin{instructionbox}{ETFR/ETTR -- Exception Link Transfer Meta-Instructions}
	\textbf{Assembles to:} \texttt{COPI \#0x1Cxy} or \texttt{COPI \#0x1Dxy}

	\textbf{Syntax:}
	\begin{lstlisting}
ETFR #n                       ; Transfer saved address and status to a and r7
ETFR a, #n                    ; Transfer saved address to a only
ETFR r7, #n                   ; Transfer saved status to r7 only
ETTR #n                       ; Transfer a and r7 to saved address and status
ETTR #n, a                    ; Transfer a to saved address only
ETTR #n, r7                   ; Transfer r7 to saved status only
\end{lstlisting}

	\textbf{Operands:} \texttt{\#n} selects an exception link register. Optional operands select whether to transfer the
	saved return address, saved status register, or both.

	\textbf{Description:} \mnemonic{ETFR} copies saved exception state from an exception link register to CPU registers.
	\mnemonic{ETTR} copies CPU register state back to the selected exception link register before returning.

	\textbf{Write-back:} \mnemonic{ETFR} writes selected saved state to \reg{a}, \reg{r7}, or both. \mnemonic{ETTR} writes
	\reg{a}, \reg{r7}, or both to selected saved exception state.

	\textbf{Flags:} Preserved by the coprocessor-call instruction.

	\textbf{Exceptions:} Raises a privilege-violation fault if executed in protected mode.

	\textbf{Condition codes:} If the condition is false, no transfer occurs and status flags are preserved.

	\textbf{Timing:} Same as \mnemonic{COPI}: 6 cycles before exception-unit transfer processing.

	\textbf{Privilege:} Supervisor mode only.

	\textbf{Example:}
	\begin{lstlisting}
alignment_fault_handler:
    ETFR a, #6                ; Get faulting return address
    LOAD al, @fixed_address   ; Correct the address
    ETTR #6, a                ; Write corrected address back
    RETE
\end{lstlisting}
\end{instructionbox}

See Chapter~\ref{ch:exceptions} for detailed documentation of exception handling and all exception coprocessor
instructions.

\section{DMA Unit (Coprocessor 2)}
\label{sec:dma-unit}

The required DMA unit provides bulk word transfers using the existing address register pairs and the general-purpose
register file. It is intentionally simple: it has no descriptor table, hidden parameter registers, or autonomous
background execution. When a DMA command is accepted, the processing unit stops fetching further instructions until the
DMA command completes or faults.

DMA operations are supervisor-only because they perform memory access using supervisor-visible address state. In
protected mode, they raise a privilege-violation fault before dispatch.

\subsection{DMA Operations}

\begin{table}[H]
	\centering
	\small
	\begin{tabularx}{\textwidth}{llllX}
		\toprule
		\textbf{Mnemonic} & \textbf{Command} & \textbf{Count}              & \textbf{Registers}                          & \textbf{Operation}                          \\
		\midrule
		\mnemonic{DMAR}   & 0x2800 | operand & \texttt{\#-7}--\texttt{\#7} & \reg{a}/\reg{l}/\reg{s}, \reg{r1}--\reg{r7} & Read memory into registers                  \\
		\mnemonic{DMAW}   & 0x2900 | operand & \texttt{\#-7}--\texttt{\#7} & \reg{a}/\reg{l}/\reg{s}, \reg{r1}--\reg{r7} & Write registers to memory                   \\
		\mnemonic{DMAT}   & 0x2A00 | n       & 0--255                      & \reg{a}, \reg{l}, \reg{r1}--\reg{r7}        & Copy memory at \reg{a} to memory at \reg{l} \\
		\bottomrule
	\end{tabularx}
	\caption{DMA Unit Operations}
	\label{tab:dma-unit-operations}
\end{table}

\noindent For \mnemonic{DMAR} and \mnemonic{DMAW}, the command operand byte is encoded as follows: bits 7--6 select the
address register (\texttt{00} = \reg{a}, \texttt{01} = \reg{l}, \texttt{10} = \reg{s}, \texttt{11} reserved), bit 5
selects direction (\texttt{0} = forward, \texttt{1} = backward), bits 2--0 hold the count magnitude, and bits 4--3 are
reserved and written as zero. Negative counts are encoded as the backward direction bit plus the positive magnitude, not
as a two's-complement count field. A zero count is encoded with the forward direction bit.

\noindent For all DMA operations, a count of zero is a no-op: no memory access occurs, registers are preserved, and
address registers are not advanced. CPU status flags are preserved. If the condition is false, no DMA command is
dispatched, no coprocessor privilege or opcode checks are performed, and all CPU state is preserved.

\noindent DMA memory reads and writes can raise data bus and bus-protection faults using DMA read/write bus access
types. If an address-register low word overflows or underflows and \texttt{SR.A} is set, the DMA operation raises a
segment-overflow fault; otherwise the low word wraps modulo $2^{16}$. Partial side effects are possible if a fault
occurs mid-transfer, and software must not assume that re-executing the same DMA instruction restarts the original
operation.

\begin{instructionbox}{DMAR -- DMA Read To Registers Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x2800 | operand}

	\textbf{Syntax:} \texttt{DMAR addr, \#n}

	\textbf{Operands:} \texttt{addr} must be \reg{a}, \reg{l}, or \reg{s}. \texttt{\#n} is a literal count operand in the
	range \texttt{\#-7}--\texttt{\#7}. Counts outside that range are invalid DMA operations.

	\textbf{Description:} Reads sequential words from the selected address register into \reg{r1} through \reg{rn}.
	Positive counts read from the current address and advance it. Negative counts first move the address register backward
	by the count magnitude, then read the ascending memory block from that new address.

	\textbf{Operation:}
	\begin{lstlisting}
count = abs(n)
if n < 0:
    addr.low = addr.low - count
transfer = addr
for i in 1..count:
    r[i] = memory[transfer]
    transfer.low = transfer.low + 1
if n > 0:
    addr.low = addr.low + count
\end{lstlisting}

	\textbf{Write-back:} A successful positive-count \mnemonic{DMAR} advances the selected address register by the count
	magnitude. A successful negative-count \mnemonic{DMAR} leaves the selected address register at its decremented start
	address. Address-register high words are not changed by the transfer increments or decrements.

	\textbf{Clobbers:} Writes \reg{r1} through \reg{rn}.

	\textbf{Flags:} Preserved.

	\textbf{Exceptions:} Unsupported DMA operations, reserved operand encodings, and invalid counts raise invalid-opcode
	faults. In protected mode, raises a privilege-violation fault. DMA reads can raise data bus, bus-protection, and
	segment-overflow faults.

	\textbf{Condition codes:} If the condition is false, no DMA command is dispatched and all CPU state is preserved.

	\textbf{Timing:} The \mnemonic{COPI} command dispatch uses the normal 6 execution phases. The DMA unit then performs
	DMA read bus cycles and may add wait states through normal bus acknowledgement.

	\textbf{Privilege:} Supervisor mode only.
\end{instructionbox}

\begin{instructionbox}{DMAW -- DMA Write From Registers Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x2900 | operand}

	\textbf{Syntax:} \texttt{DMAW addr, \#n}

	\textbf{Operands:} \texttt{addr} must be \reg{a}, \reg{l}, or \reg{s}. \texttt{\#n} is a literal count operand in the
	range \texttt{\#-7}--\texttt{\#7}. Counts outside that range are invalid DMA operations.

	\textbf{Description:} Writes \reg{r1} through \reg{rn} to sequential words at the selected address register. Positive
	counts write from the current address and advance it. Negative counts first move the address register backward by the
	count magnitude, then write the ascending memory block from that new address. Register order always corresponds to
	ascending memory addresses, so \texttt{DMAW s, \#-n} followed by \texttt{DMAR s, \#n} restores \reg{r1} through
	\reg{rn}.

	\textbf{Operation:}
	\begin{lstlisting}
count = abs(n)
if n < 0:
    addr.low = addr.low - count
transfer = addr
for i in 1..count:
    memory[transfer] = r[i]
    transfer.low = transfer.low + 1
if n > 0:
    addr.low = addr.low + count
\end{lstlisting}

	\textbf{Write-back:} A successful positive-count \mnemonic{DMAW} advances the selected address register by the count
	magnitude. A successful negative-count \mnemonic{DMAW} leaves the selected address register at its decremented start
	address. Address-register high words are not changed by the transfer increments or decrements.

	\textbf{Clobbers:} General-purpose registers are preserved.

	\textbf{Flags:} Preserved.

	\textbf{Exceptions:} Unsupported DMA operations, reserved operand encodings, and invalid counts raise invalid-opcode
	faults. In protected mode, raises a privilege-violation fault. DMA writes can raise data bus, bus-protection, and
	segment-overflow faults.

	\textbf{Condition codes:} If the condition is false, no DMA command is dispatched and all CPU state is preserved.

	\textbf{Timing:} The \mnemonic{COPI} command dispatch uses the normal 6 execution phases. The DMA unit then performs
	DMA write bus cycles and may add wait states through normal bus acknowledgement.

	\textbf{Privilege:} Supervisor mode only.
\end{instructionbox}

\begin{instructionbox}{DMAT -- DMA Memory Transfer Meta-Instruction}
	\textbf{Assembles to:} \texttt{COPI \#0x2A00 | n}

	\textbf{Syntax:} \texttt{DMAT a, l, \#n}

	\textbf{Operands:} The source and destination address registers must be written explicitly as \reg{a} and \reg{l}.
	\texttt{\#n} is an 8-bit literal count operand in the range 0--255. A count of zero is a no-op.

	\textbf{Description:} Copies sequential words from the address in \reg{a} to the address in \reg{l}, using
	\reg{r1}--\reg{r7} as an internal transfer window in chunks of up to seven words.

	\textbf{Operation:}
	\begin{lstlisting}
while remaining > 0:
    chunk = min(remaining, 7)
    read chunk words from memory[a] into r1..r[chunk]
    write chunk words from r1..r[chunk] to memory[l]
    a.low = a.low + chunk
    l.low = l.low + chunk
    remaining = remaining - chunk
\end{lstlisting}

	\textbf{Write-back:} A successful non-zero \mnemonic{DMAT} advances \reg{a} and \reg{l} by the number of words
	transferred. Address-register high words are not changed by the transfer increments.

	\textbf{Clobbers:} Clobbers \reg{r1}--\reg{r7}; after completion they contain the final transfer window.

	\textbf{Flags:} Preserved.

	\textbf{Exceptions:} Unsupported DMA operations and reserved operand encodings raise invalid-opcode faults. In
	protected mode, raises a privilege-violation fault. DMA reads and writes can raise data bus, bus-protection, and
	segment-overflow faults.

	\textbf{Fault side effects:} Registers, memory, and address registers reflect the last completed word transfer if a
	fault occurs mid-transfer.

	\textbf{Condition codes:} If the condition is false, no DMA command is dispatched and all CPU state is preserved.

	\textbf{Timing:} The \mnemonic{COPI} command dispatch uses the normal 6 execution phases. The DMA unit then performs
	DMA read and write bus cycles and may add wait states through normal bus acknowledgement. Instruction fetch resumes
	only after the DMA operation completes or faults.

	\textbf{Privilege:} Supervisor mode only.

	\textbf{Overlap:} Overlapping source and destination ranges are architecturally undefined unless the source and
	destination addresses are identical.
\end{instructionbox}

\section{Integer Maths Unit (Coprocessor 3)}
\label{sec:integer-maths-unit}

The optional integer maths unit provides a stable multiply/divide convention. CPU models without the hardware raise an
invalid-opcode fault for these commands, allowing supervisor software to emulate the operation and return as if the
coprocessor call had completed.

Reference source listings for software emulation routines are supplied with the SIRC-1 distribution media. Those
listings are examples of a supervisor compatibility handler, not additional instruction encodings.

\subsection{Maths Operations}

\begin{table}[H]
	\centering
	\small
	\begin{tabularx}{\textwidth}{llllX}
		\toprule
		\textbf{Mnemonic} & \textbf{Command} & \textbf{Inputs}             & \textbf{Outputs}             & \textbf{Operation}              \\
		\midrule
		\mnemonic{MULU}   & 0x3000           & \reg{r1}, \reg{r2}          & \reg{r1}:\reg{r2}, \reg{r3}  & Unsigned 16 x 16 -> 32 multiply \\
		\mnemonic{MULS}   & 0x3100           & \reg{r1}, \reg{r2}          & \reg{r1}:\reg{r2}, \reg{r3}  & Signed 16 x 16 -> 32 multiply   \\
		\mnemonic{DIVU}   & 0x3200           & \reg{r1}:\reg{r2}, \reg{r3} & \reg{r1}, \reg{r2}, \reg{r3} & Unsigned divide                 \\
		\mnemonic{DIVS}   & 0x3300           & \reg{r1}:\reg{r2}, \reg{r3} & \reg{r1}, \reg{r2}, \reg{r3} & Signed divide                   \\
		\bottomrule
	\end{tabularx}
	\caption{Integer Maths Unit Operations}
	\label{tab:integer-maths-unit-operations}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{cll}
		\toprule
		\textbf{Bit} & \textbf{Name} & \textbf{Meaning}             \\
		\midrule
		0            & E             & Any maths error              \\
		1            & DZ            & Divide by zero               \\
		2            & OV            & Quotient overflow            \\
		3--15        & Reserved      & Written as zero, ignore read \\
		\bottomrule
	\end{tabular}
	\caption{Maths Status Word in \reg{r3}}
	\label{tab:maths-status-word}
\end{table}

\noindent Integer maths operations preserve CPU status flags. Missing maths hardware and unsupported maths operations
raise invalid-opcode faults. Divide by zero and quotient overflow are reported in \reg{r3}, not as arithmetic
exceptions. If the condition is false, no maths command is dispatched, no coprocessor opcode check is performed, and
registers and status flags are preserved.

\begin{instructionbox}{MULU/MULS -- Integer Multiply Meta-Instructions}
	\textbf{Assembles to:} \texttt{COPI \#0x3000} or \texttt{COPI \#0x3100}

	\textbf{Syntax:}
	\begin{lstlisting}
MULU                          ; Unsigned multiply
MULS                          ; Signed multiply
\end{lstlisting}

	\textbf{Operands:} None in assembly syntax. The operations multiply \reg{r1} by \reg{r2}.

	\textbf{Description:} Multiplies two 16-bit operands and writes the 32-bit product to \reg{r1}:\reg{r2}, high word
	first. \mnemonic{MULU} treats both inputs as unsigned. \mnemonic{MULS} treats both inputs as two's-complement signed
	values.

	\textbf{Operation:}
	\begin{lstlisting}
MULU:
    product = unsigned(r1) * unsigned(r2)
    r1 = product[31:16]
    r2 = product[15:0]
    r3 = 0

MULS:
    product = signed(r1) * signed(r2)
    r1 = product[31:16]
    r2 = product[15:0]
    r3 = 0
\end{lstlisting}

	\textbf{Write-back:} Writes \reg{r1}, \reg{r2}, and \reg{r3}.

	\textbf{Flags:} CPU status flags are preserved. \reg{r3} is written as zero.

	\textbf{Exceptions:} Missing maths hardware and unsupported multiply operations raise invalid-opcode faults. These
	operations perform no data-memory bus access.

	\textbf{Condition codes:} If the condition is false, no maths command is dispatched and registers and status flags are
	preserved.

	\textbf{Timing:} The \mnemonic{COPI} command dispatch uses the normal 6 execution phases. Hardware maths operations
	may take additional implementation-defined cycles before instruction fetch resumes.

	\textbf{Privilege:} User-callable in protected mode and supervisor mode.
\end{instructionbox}

\begin{instructionbox}{DIVU/DIVS -- Integer Divide Meta-Instructions}
	\textbf{Assembles to:} \texttt{COPI \#0x3200} or \texttt{COPI \#0x3300}

	\textbf{Syntax:}
	\begin{lstlisting}
DIVU                          ; Unsigned divide
DIVS                          ; Signed divide
\end{lstlisting}

	\textbf{Operands:} None in assembly syntax. The dividend is \reg{r1}:\reg{r2}, high word first. The divisor is
	\reg{r3}.

	\textbf{Description:} Divides a 32-bit dividend by a 16-bit divisor. Successful division writes the remainder to
	\reg{r1}, the quotient to \reg{r2}, and zero to \reg{r3}.

	\textbf{Operation:}
	\begin{lstlisting}
if divisor is zero:
    r3 = E | DZ
else if quotient does not fit in 16 bits:
    r3 = E | OV
else:
    r1 = remainder
    r2 = quotient
    r3 = 0
\end{lstlisting}

	\textbf{Write-back:} Successful division writes \reg{r1}, \reg{r2}, and \reg{r3}. Divide-by-zero and
	quotient-overflow cases leave \reg{r1}:\reg{r2} unchanged and write the maths status word to \reg{r3}.

	\textbf{Flags:} CPU status flags are preserved. Maths errors are reported through \reg{r3}, not through \reg{sr}.

	\textbf{Exceptions:} Missing maths hardware and unsupported divide operations raise invalid-opcode faults. Divide by
	zero and quotient overflow do not raise arithmetic exceptions. These operations perform no data-memory bus access.

	\textbf{Condition codes:} If the condition is false, no maths command is dispatched and registers and status flags are
	preserved.

	\textbf{Timing:} The \mnemonic{COPI} command dispatch uses the normal 6 execution phases. Hardware maths operations
	may take additional implementation-defined cycles before instruction fetch resumes. Software emulation timing is
	defined by the invalid-opcode handler.

	\textbf{Privilege:} User-callable in protected mode and supervisor mode.

	\textbf{Signed division:} \mnemonic{DIVS} treats \reg{r1}:\reg{r2} as a 32-bit two's-complement dividend and
	\reg{r3} as a 16-bit two's-complement divisor. The quotient is truncated toward zero and the remainder has the sign
	of the dividend. The case \texttt{-2147483648 / -1} is quotient overflow.
\end{instructionbox}
