\chapter{Meta-Instructions}
\label{ch:meta-instructions}

\section{Overview}

Meta-instructions (also called pseudo-instructions) are assembly mnemonics that don't directly correspond to unique
opcodes. Instead, they assemble to specific forms of existing instructions, providing more intuitive syntax for common
operations.

\section{NOOP -- No Operation}

\textbf{Syntax:} \texttt{NOOP}

\textbf{Assembles to:} \texttt{ADDI r1, \#0}

\textbf{Description:} Does nothing for 6 cycles. Useful for timing delays, instruction padding, or placeholders during development.

\textbf{Example:}
\begin{lstlisting}
NOOP                          ; Wait 6 cycles
NOOP
NOOP                          ; Total 18 cycles delay
\end{lstlisting}

\section{RETS -- Return from Subroutine}

\textbf{Syntax:} \texttt{RETS}

\textbf{Assembles to:} \texttt{LDEA p, (\#0, l)}

\textbf{Description:} Returns from a subroutine by copying the link register to the program counter. Used after \mnemonic{LJSR} or \mnemonic{BRSR}.

\textbf{Example:}
\begin{lstlisting}
my_function:
    ; ... function body ...
    RETS                      ; Return to caller
\end{lstlisting}

\section{LJMP -- Long Jump}

\textbf{Syntax:} \texttt{LJMP src [, offset]}

\textbf{Assembles to:} \texttt{LDEA p, ...} (various forms)

\textbf{Description:} Performs an absolute jump by transferring an address register to the program counter. Supports optional immediate or register offsets. Unlike \mnemonic{BRAN}, which is PC-relative, \mnemonic{LJMP} uses absolute addressing via address registers.

\textbf{Examples:}
\begin{lstlisting}
; Simple jump to address in register
LJMP a                        ; Assembles to: LDEA p, (#0, a)

; Jump with immediate offset
LJMP a, #16                   ; Assembles to: LDEA p, (#16, a)

; Jump with register displacement
LJMP a, r1                    ; Assembles to: LDEA p, (r1, a)
\end{lstlisting}

\section{LJSR -- Long Jump to Subroutine}

\textbf{Syntax:} \texttt{LJSR src [, offset]}, \texttt{LJSR (\#offset, src)+}, or \texttt{LJSR (rS, src)+}

\textbf{Assembles to:} \texttt{LDEL p, ...} (various forms)

\textbf{Description:} Performs an absolute call through an address register. The return address is saved to \reg{l}; the
computed target address is written to \reg{p}. Supports optional immediate or register offsets, plus post-increment
forms that advance the source address register after computing the call target.

\textbf{Examples:}
\begin{lstlisting}
; Simple call to address in register
LJSR a                        ; Assembles to: LDEL p, (#0, a)

; Call with immediate offset
LJSR a, #16                   ; Assembles to: LDEL p, (#16, a)

; Call with register displacement
LJSR a, r1                    ; Assembles to: LDEL p, (r1, a)

; Call and post-increment the source pointer
LJSR (#0, a)+                 ; Assembles to: LDEL p, (#0, a)+
LJSR (r1, a)+                 ; Assembles to: LDEL p, (r1, a)+
\end{lstlisting}

\section{BRAN -- PC-Relative Branch}

\textbf{Syntax:} \texttt{BRAN \#disp} or \texttt{BRAN @label}

\textbf{Assembles to:} \texttt{LDEA p, (\#disp, p)}

\textbf{Description:} Branches relative to the current program counter. Label references are resolved by the linker as
PC-relative offsets.

\section{BRSR -- PC-Relative Branch to Subroutine}

\textbf{Syntax:} \texttt{BRSR \#disp} or \texttt{BRSR @label}

\textbf{Assembles to:} \texttt{LDEL p, (\#disp, p)}

\textbf{Description:} Calls a PC-relative subroutine, saving the return address to \reg{l}. Label references are resolved
by the linker as PC-relative offsets.

\section{EXCP -- User Exception (Trap)}

\textbf{Syntax:} \texttt{EXCP \#vector}

\textbf{Assembles to:} \texttt{COPI \#0x1100 | vector}

\textbf{Description:} Triggers a software exception (trap) at the specified user vector (0x60-0xFF). Used by user-mode programs to invoke system calls or request supervisor mode services.

\textbf{Example:}
\begin{lstlisting}
; System call to print character
LOAD r1, #'A'                 ; Character to print
EXCP #0x80                    ; Trap to OS print routine
\end{lstlisting}

\section{WAIT -- Wait for Interrupt}

\textbf{Syntax:} \texttt{WAIT}

\textbf{Assembles to:} \texttt{COPI r1, \#0x1900}

\textbf{Description:} Puts the CPU into a low-power wait state until an exception occurs. Sets the waiting\_for\_exception flag in the exception unit. Only valid in supervisor mode.

\textbf{Example:}
\begin{lstlisting}
main_loop:
    WAIT                      ; Sleep until interrupt
    ; ... process interrupt ...
    BRAN @main_loop
\end{lstlisting}

\section{RETE -- Return from Exception}

\textbf{Syntax:} \texttt{RETE}

\textbf{Assembles to:} \texttt{COPI \#0x1A00}

\textbf{Description:} Returns from an exception handler by restoring the saved program counter and status register. Only valid in supervisor mode.

\textbf{Example:}
\begin{lstlisting}
exception_handler:
    ; ... handle exception ...
    RETE                      ; Return to interrupted code
\end{lstlisting}

\section{RSET -- System Reset}

\textbf{Syntax:} \texttt{RSET}

\textbf{Assembles to:} \texttt{COPI \#0x1B00}

\textbf{Description:} Performs a software reset of the processor by clearing the status register and jumping to the reset vector. Only valid in supervisor mode.

\textbf{Example:}
\begin{lstlisting}
fatal_error:
    ; Log error...
    RSET                      ; Reset the system
\end{lstlisting}

\section{ETFR -- Exception Transfer From Register}

\textbf{Syntax:}
\begin{lstlisting}
ETFR #n                       ; Transfer both to a and r7
ETFR a, #n                    ; Transfer address to a only
ETFR r7, #n                   ; Transfer status to r7 only
\end{lstlisting}

\textbf{Assembles to:} \texttt{COPI \#0x1Cxy} (where x = register select, y = link register)

\textbf{Description:} Copies the return address and/or status register from a link register (0-7) to CPU registers. Used in exception handlers to examine saved processor state. Only valid in supervisor mode.

\textbf{Example:}
\begin{lstlisting}
alignment_fault_handler:
    ETFR a, #6                ; Get faulting return address
    LOAD al, @fixed_address   ; Correct the address
    ETTR #6, a                ; Write corrected address back
    RETE
\end{lstlisting}

\section{ETTR -- Exception Transfer To Register}

\textbf{Syntax:}
\begin{lstlisting}
ETTR #n                       ; Transfer both a and r7
ETTR #n, a                    ; Transfer a to address only
ETTR #n, r7                   ; Transfer r7 to status only
\end{lstlisting}

\textbf{Assembles to:} \texttt{COPI \#0x1Dxy} (where x = register select, y = link register)

\textbf{Description:} Copies CPU registers to a link register's return address and/or status register. Used to modify saved processor state before returning from an exception. Only valid in supervisor mode.

\textbf{Example:}
\begin{lstlisting}
privilege_fault_handler:
    ETFR r7, #6               ; Get saved status register
    ANDI r7, #0xFEFF          ; Clear protected mode bit
    ETTR #6, r7               ; Save modified status back
    RETE
\end{lstlisting}

\section{SHFT -- Shift with Status Update}

\textbf{Syntax:} \texttt{SHFT rD, shift}

\textbf{Assembles to:} \texttt{ORRI[S] rD, \#0, shift}

\textbf{Description:} Provides concise syntax for shifting a register value while taking status flags from the shifter
result. \mnemonic{SHFT} is not a separate CPU operation. It assembles to a short-immediate \mnemonic{ORRI} form with an
immediate value of zero and the status-update source set to \texttt{[S]}. The zero OR operation preserves the shifted
source value, while \texttt{[S]} makes the status flags reflect the shift result instead of the OR result.

Normally, shifted ALU instructions such as \texttt{CMPI r2, \#1, LSL \#2} update status flags from the ALU result
unless an explicit status source override is present. \mnemonic{SHFT} is the assembler convenience for the common case
where the programmer wants the shifted register value and flags from that same shifted value.

\textbf{Examples:}
\begin{lstlisting}
; Shift r1 left by 3 and update flags
SHFT r1, LSL #3               ; ORRI[S] r1, #0, LSL #3

; Shift right and test result
SHFT r3, LSR #2               ; r3 = r3 >> 2
BRAN|== @zero_result          ; Branch if shifted result is zero

; Arithmetic shift with sign check
SHFT r6, ASR #4               ; r6 = r6 >> 4 (signed)
BRAN|NS @negative_value       ; Branch if result is negative
\end{lstlisting}

\textbf{Notes:}
\begin{itemize}
	\item The underlying status override syntax is described in Section~\ref{sec:status-override}.
	\item All shift types are supported (LSL, LSR, ASL, ASR, RTL, RTR)
	\item Both immediate and register-based shift counts are supported
	\item Because the lowering uses the short-immediate format, \mnemonic{SHFT} has the same shift field limits as other
	      short-immediate ALU instructions.
\end{itemize}
