\chapter{Instruction Timing}
\label{appendix:timing}

\section{Overview}

All SIRCIS instructions follow the six-stage execution phase model and complete in 6 clock cycles when no bus wait
states are required. A bus operation may hold the current phase until it is acknowledged, adding wait cycles without
adding execution phases. This fixed execution sequence greatly simplifies software development and hardware design.

\section{Six-Stage Execution Phases}

For a detailed description of each phase, see Chapter~\ref{ch:cpu-architecture}.

\begin{enumerate}
	\item \textbf{Instruction Fetch (High Word)} -- 1 cycle
	\item \textbf{Instruction Fetch (Low Word)} -- 1 cycle
	\item \textbf{Decode and Register Fetch} -- 1 cycle
	\item \textbf{Execute and Address Calculation} -- 1 cycle
	\item \textbf{Memory Access} -- 1 cycle
	\item \textbf{Write Back} -- 1 cycle
\end{enumerate}

\section{Timing Characteristics}

\begin{table}[H]
	\centering
	\begin{tabular}{lc}
		\toprule
		\textbf{Instruction Type}  & \textbf{Cycles} \\
		\midrule
		ALU (Register)             & 6               \\
		ALU (Immediate)            & 6               \\
		Load from Memory           & 6               \\
		Store to Memory            & 6               \\
		Branch (taken)             & 6               \\
		Branch (not taken)         & 6               \\
		Conditional (executed)     & 6               \\
		Conditional (not executed) & 6               \\
		NOOP                       & 6               \\
		\bottomrule
	\end{tabular}
	\caption{Instruction Timing (All instructions)}
\end{table}

The cycle counts in this appendix assume that every bus operation is acknowledged without a wait state.

\section{Conditional Execution}

When a conditional instruction's condition is not met:
\begin{itemize}
	\item The instruction still takes 6 cycles to complete when no bus wait states are required
	\item All execution phases run normally
	\item The write-back stage is disabled (no register or memory modification)
	\item Status flags are not updated
\end{itemize}

This means conditional instructions do not branch and avoid control-flow stalls, but failed conditions still consume
execution time.

\section{Program Timing Examples}

\subsection{Simple Loop}
\begin{lstlisting}
; Process 10 elements (assuming no bus wait states)
ADDI r7, #0, #10              ; 6 cycles
loop:
    LOAD r1, (#0, a)+         ; 6 cycles
    ADDI r1, #1               ; 6 cycles
    STOR (#-2, a), r1         ; 6 cycles
    SUBI r7, #1               ; 6 cycles
    BRAN|!= (#loop, p)        ; 6 cycles
; Total per iteration: 30 cycles
; Total for 10 iterations: 6 + (10 * 30) = 306 cycles
\end{lstlisting}

\subsection{Function Call Overhead}
\begin{lstlisting}
; Call function
LJSR (#0, a)                  ; 6 cycles

; Function body executes...

; Return
RETS                          ; 6 cycles
; Total overhead: 12 cycles
\end{lstlisting}

\subsection{Conditional Block}
\begin{lstlisting}
CMPR r1, r2                   ; 6 cycles
ADDR|HI r3, r4, r5            ; 6 cycles (exec if r1 > r2)
ADDR|LO r3, r6, r7            ; 6 cycles (exec if r1 <= r2)
; Total: 18 cycles (one conditional executes, one doesn't)
\end{lstlisting}

\section{Performance Considerations}

\subsection{No Pipeline Stalls}

The SIRC-1 does not implement pipelining in the traditional sense where multiple instructions are in flight
simultaneously. Instead:

\begin{itemize}
	\item Each instruction completes fully before the next begins
	\item No data hazards or phase bubbles
	\item No branch prediction penalties
	\item Base execution time is predictable before bus wait states are considered
\end{itemize}

\subsection{Memory Access}

\begin{itemize}
	\item Memory access occupies one execution phase
	\item If BACK is not asserted, the CPU holds that phase until the bus operation is acknowledged
	\item Each wait state adds one clock cycle to the instruction
\end{itemize}

\subsection{Optimization Guidelines}

\begin{enumerate}
	\item \textbf{Use conditional execution for short sequences}
	      \begin{lstlisting}
; Instead of:
CMPR r1, r2
BRAN|<= skip
ADDI r3, #1
skip:

; Use:
CMPR r1, r2
ADDI|HI r3, #1             ; Saves branch overhead
    \end{lstlisting}

	\item \textbf{Combine operations with shifts}
	      \begin{lstlisting}
; Instead of:
ADDR r1, r2, r2
ADDI r1, #0, LSL #1

; Use:
ADDR r1, r2, r2, LSL #1    ; Multiply by 3 in one instruction
    \end{lstlisting}

	\item \textbf{Use post-increment/pre-decrement for iteration}
	      \begin{lstlisting}
; Instead of:
LOAD r1, (#0, a)
ADDI a, #2

; Use:
LOAD r1, (#0, a)+          ; Same result, fewer cycles
    \end{lstlisting}

	\item \textbf{Loop unrolling reduces overhead}
	      \begin{lstlisting}
; Instead of processing one item per iteration,
; process multiple items to reduce branch overhead
loop:
    LOAD r1, (#0, a)+
    LOAD r2, (#0, a)+
    LOAD r3, (#0, a)+
    LOAD r4, (#0, a)+
    ; Process 4 items...
    SUBI r7, #4
    BRAN|HI loop
    \end{lstlisting}
\end{enumerate}

\section{Theoretical Performance}

At a given clock frequency:

\begin{itemize}
	\item \textbf{Instructions per second:} $f_{clock} / 6$
	\item \textbf{At 10 MHz:} $\approx$ 1.67 million instructions/second
	\item \textbf{At 25 MHz:} $\approx$ 4.17 million instructions/second
\end{itemize}

These are theoretical maximums assuming perfect code with no inefficiencies.

\section{Comparison to Other CPUs}

\begin{table}[H]
	\centering
	\small
	\begin{tabular}{lcp{6cm}}
		\toprule
		\textbf{CPU}   & \textbf{Avg CPI} & \textbf{Notes}             \\
		\midrule
		SIRC-1         & 6.0              & Fixed, all instructions    \\
		MOS 6502       & 2-7              & Variable by instruction    \\
		Motorola 68000 & 4-158            & Highly variable            \\
		ARM6           & 1.0              & Single-cycle with pipeline \\
		MIPS R2000     & 1.0              & Single-cycle with pipeline \\
		\bottomrule
	\end{tabular}
	\caption{Cycles Per Instruction Comparison}
\end{table}

The SIRC-1's fixed 6-cycle execution makes it slower than pipelined RISC CPUs but more predictable. It's comparable to
early microprocessors while being simpler to implement in hardware.
